---
title: "Community analysis"
runtime: shiny
output: 
  html_document:
    toc: true
    toc_float: 
       collapsed: false
    number_sections: true
---

```{r setupcomman, include = FALSE}
# globals
library(knitr)
opts_chunk$set(echo = F, message = F, warning = F)

library(tidyverse)
library(sf)
library(mapview)
library(shiny)
library(scales)
library(leaflet)
library(ggord)
library(vegan)
library(ggdendro)
library(dendextend)
library(ggrepel)

# data, funcs, globals
prj <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"

source('R/funcs.R')

data(fishdat)
data(stream)

# map colors
cols <- mapviewGetOption("vector.palette")(4)

colgrp <- data.frame(
  Watershed = c('APT', 'PAJ', 'SLR', 'SOQ'), 
  cols = cols, 
  stringsAsFactors = F
)

# ggplot base 
pbase <- theme_bw(base_family = 'serif') +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10), 
    axis.text.y = element_text(size = 10),
    legend.position = 'top',
    legend.direction = 'horizontal',
    # plot.margin = unit(c(4,4,0,0), "lines"),
    strip.background = element_blank(), 
    strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5), 
    panel.background = element_rect(fill = 'black')
  ) 

# prep species data, long format 
spdat <- fishdat %>% 
  dplyr::select(Year, Watershed, SiteID, matches('^Sp_')) %>% 
  gather('species', 'pa', -Year, -SiteID, -Watershed, -geometry) %>% 
  mutate(
    species = gsub('^Sp_', '', species)
  ) %>% 
  filter(!is.na(Year)) %>% 
  group_by(Year, species) %>% 
  filter(!duplicated(SiteID)) %>% 
  ungroup() 

# site, wshed lookup
siteshd <- fishdat
st_geometry(siteshd) <- NULL
siteshd <- siteshd %>% 
  dplyr::select(SiteID, Watershed) %>% 
  unique %>% 
  mutate_if(is.factor, as.character)
```

<br>

These analyses can be used to evaluate changes in community composition over time using the presence/absence data for all species.  The first two sections show general changes by species and the third section is a multivariate analysis showing ordination and clustering results of the communities for a selected year.  Changes over time can be assessed by comparing the results between years. 

# Percent occupied sites over time

These plots show the percent of sampled sites over time that are occupied with a selected species, grouped by watershed.  __The total number of sites that were sampled in each watershed, year combination is shown inside the point.__

```{r tmsel}
column(2, NULL)
# select a species
column(width = 4,
 selectInput('sptm', 'Select a species:', sort(unique(spdat$species)), selected = 'Sthd')
)

# year slider
column(width = 4, 
  sliderInput("yrs", label = 'Select year ranges to plot:',  
        min = 1981, max = 2018, 
        value = c(1994, 2018),
        sep = '', ticks = FALSE
      )
)
column(2, NULL)
```

```{r spyragg}
# percent of sites where species was observed, by watershed, year
spyragg <- reactive({
  
  # inputs
  sp <- input$sptm
  yrs <- input$yrs
  
  # tabulate percent sites occupied
  out <- spdat
  st_geometry(out) <- NULL
  out <- out %>% 
    filter(species %in% sp) %>% 
    filter(Year >= yrs[1] & Year <= yrs[2]) %>% 
    group_by(Watershed, Year) %>% 
    summarize(
      nprs = sum(pa),
      nsites = n()
    ) %>% 
    mutate(
      prcsit = 100 * nprs / nsites
    )
  
  return(out)
  
})

# relative occurrence of a species at each site by year selection
spyrsocc <- reactive({
  
  # inputs
  sp <- input$sptm2
  yrs <- input$yrs2

  # get species, year selection
  toplo <- spdat %>% 
    filter(species %in% sp) %>% 
    filter(Year >= yrs[1] & Year <= yrs[2]) 
 
  # get geometry for later join
  spgeo <- toplo %>% 
    dplyr::select(SiteID) %>% 
    filter(!duplicated(SiteID))

  # remove geometry for aggregate, get cats
  st_geometry(toplo) <- NULL
  toplo <- toplo %>%  
    unique %>% 
    group_by(SiteID) %>% 
    summarize(
      nprs = sum(pa),
      nsites = n()
    ) %>% 
    mutate(
      prcsit = 100 * nprs / nsites, 
      catsit = cut(prcsit, breaks = c(-Inf, 0, 25, 50, 75, Inf), labels = c('absent', 'rare', 'occasional', 'frequent', 'common')),
      catcol = factor(catsit, levels = c('absent', 'rare', 'occasional', 'frequent', 'common'), labels = c("#FFFFFF", "#E31A1C", "#FB9A99", "#B2DF8A", "#33A02C")), 
      catcol = as.character(catcol)
    ) %>% 
    left_join(spgeo, ., by = 'SiteID')
  
  # colors must be passed as vector
  cols <- toplo$catcol
  
  out <- mapview(stream, label = stream$STREAM_NM, homebutton = F, legend = F) +
    mapview(toplo, zcol = 'catsit', cex = 5, label = paste0(toplo$SiteID, ': ', toplo$catsit), legend = F, homebutton = F, col.region = cols)
  
  return(out)
  
})
```

```{r prsplo, out.width = "100%"}
renderPlot({
  
  # inputs
  toplo <- spyragg()
  sptm <- input$sptm
  
  p <- ggplot(toplo, aes(y = prcsit, x = Year)) + 
    geom_line() + 
    geom_point(pch = 21, aes(fill = prcsit), size = 8) +
    facet_wrap(~Watershed, ncol = 2) + 
    theme_bw(base_family = 'serif', base_size = 18) +
    theme(
      strip.background = element_blank(),
      axis.title.x = element_blank(), 
      legend.position = 'none'
    ) + 
    scale_y_continuous(paste0('% sites with ', sptm), limits = c(0, 100)) + 
    geom_text(aes(label = nsites), vjust = 0.5, size = 4, color = 'black') + 
    scale_fill_gradientn(colors = c('tomato1', 'lightgreen'), limits = c(0, 100))

  return(p)
                       
},width = 900, height = 600)

```

# Presence/absence by year

This map shows species presence/absence for a selected species and year.  Sites that were sampled for the selected year are shown on the map with small red points indicating a selected species was not found and large green points indicating a species was found at a site.  The year selector can be toggled to show only years when a species was present at any site versus all years for which sampling occurred but a species may or may not have been found (default). 

```{r sppinput}
column(12, 
      column(2, NULL),
      # select a species
      column(width = 4,
       selectInput('sp', 'Select a species:', sort(unique(spdat$species)), selected = 'Sthd')
      ), 
      
      column(width = 4, 
             selectInput('yrsub', 'Select from years when present:', selected = FALSE, choices = c(TRUE, FALSE))
      ),
      column(2, NULL)
)

column(12, 
       
      column(2, NULL),
      
      column(4,
      renderUI({
        
        # inputs
        yrsub <- input$yrsub
        mapsp <- mapsp()

        # get only years when species was found, otherwise all
        if(yrsub){

          yrs <- mapsp %>%  
            filter(pa %in% 1) %>% 
            pull(Year) %>% 
            unique

        } else {

          yrs <- unique(mapsp$Year)

        }

        # select a year
        selectInput('yr', 'Select a year:', selected = max(yrs), choices = yrs)
               
      })
      
      ), 
      
      column(6, NULL)
      
)
```

```{r prepmapsp}
# get selected species data for reactive UI year selection
mapsp <- reactive({
  
  sp <- input$sp
  
  out <- spdat %>% 
    filter(species == sp) 
  
  return(out)
  
})

# subset year and species,  map
mapspyr <- reactive({
  
  yr <- input$yr
  mapsp <- mapsp()
  
  toplo <- mapsp %>% 
    filter(Year == yr) %>% 
    mutate(
      cexv = ifelse(pa == 1, 7, 3), 
      pa = factor(pa, levels = c('1', '0'), labels = c('present', 'absent')), 
      cols = ifelse(pa == 'present', 'lightgreen', 'tomato1')
      )
  
  # colors must be passed as vector
  cols <- toplo$cols
  
  out <- mapview(stream, label = stream$STREAM_NM, homebutton = F, legend = F) +
    mapview(toplo, zcol = 'pa', cex = toplo$cexv, label = paste0(toplo$SiteID, ': ', toplo$pa), legend = F, homebutton = F, col.region = cols)
   
  return(out)
  
})
```

```{r mapsp}
renderLeaflet({mapspyr()@map})
```
<br>

# Frequency occurrence over time

This map shows the occurrence of `r renderText({input$sptm})` at each site from `r renderText({input$yrs2[1]})` to `r renderText({input$yrs2[2]})`.  The colors indicate if a species was commonly (dark green), frequently (light green), occasionally (light red), or rarely (dark red) observed at a site.  Frequency occurrence was the sum of years for which a species was observed divided by the total number of years that a site was sampled.

<span style="color:#33A02C">Common</span>: 76 - 100% of all sampled years 

<span style="color:#B2DF8A">Frequently present</span>: 51 - 75% of all sampled years 

<span style="color:#FB9A99">Occasionally present</span>: 26 - 50% of all sampled years 

<span style="color:#E31A1C">Rare</span>: >0 - 25% of all sampled years 

No fill: Absent

```{r tmsel2}
column(2, NULL)
# select a species
column(width = 4,
 selectInput('sptm2', 'Select a species:', sort(unique(spdat$species)), selected = 'Sthd')
)

# year slider
column(width = 4, 
  sliderInput("yrs2", label = 'Select year ranges to plot:',  
        min = 1981, max = 2018, 
        value = c(1994, 2018),
        sep = '', ticks = FALSE
      )
)
column(2, NULL)
```

```{r}
renderLeaflet({spyrsocc()@map})
```

<br>

# Multivariate community analysis {.tabset}

This section can be used to evaluate community changes over time by showing natural groupings between sites using species presence/absence data.  Some questions to consider when evaluating the data are:

1) Do the results differ when evaluating the results between years?
2) Can the results be explained by community differences between the watersheds?
3) If not, are there other natural groupings in the data that can describe community differences?


```{r mltsetup}
mltdat <- spdat
st_geometry(mltdat) <- NULL

column(12, 

       column(2, NULL),
       
       # select a year 
       column(width = 4, 
              selectInput('yrmlt1', 'Select a year for comparison:', sort(unique(mltdat$Year)), selected = 2018)
       ),
       
       # select cluster cuts
       column(width = 4, 
              selectInput('clstcut', 'Select cluster groups:', selected = 2, choices = c(1:10))
              ),
       
       column(2, NULL)
       
)
```

<br>

```{r ord1}
# data for ordination first year
mltdatyr1 <- reactive({
  
  # inputs
  yrmlt1 <- input$yrmlt1
  remrr <- 1 

  # create species pa mat after filtering species at less than remrr sites
  toord <- mltdat %>% 
    filter(Year == yrmlt1) %>% 
    dplyr::select(SiteID, Watershed, species, pa) %>% 
    group_by(species) %>% 
    mutate(totsp = sum(pa)) %>% 
    filter(totsp > remrr) %>% 
    dplyr::select(-totsp) %>% 
    spread(species, pa) %>% 
    data.frame(stringsAsFactors = F)
  rownames(toord) <- NULL
  toord <- toord %>% 
    column_to_rownames('SiteID')

  # remove sites with nothing found after rare spp removed
  torm <- rowSums(toord[, !names(toord) %in% c('Watershed')])
  toord <- toord[torm > 0, ]
  
  return(toord)
  
})

# wshed groups for ggord
wshdord1 <- reactive({
  
  mltdatyr1() %>% pull(Watershed)
  
})

# dismat year 1
dis1 <- reactive({
  
  # input
  distyp <- 'jaccard'
  
  toord <- mltdatyr1() %>% 
    dplyr::select(-Watershed) 
  
  # metaMDS
  dis <- toord %>% 
    vegdist(method = distyp)
  
  return(dis)
  
})

# ordination year 1
ord1 <- reactive({
  
  # input
  distyp <- 'jaccard'
  
  toord <- mltdatyr1() %>% 
    dplyr::select(-Watershed) 
  
  # metaMDS
  ord <- toord %>% 
    metaMDS(distance = distyp)
  
  return(ord)
  
})
```

## Biplot

```{r}
# some chatty stuff about the ordination
renderText({
  
  # input
  toprnt <- ord1()
  
  strs <- toprnt %>% .$stress %>% round(2)
  cnvr <- toprnt %>% .$converged
  
  if(cnvr)
    txt <- paste0('Ordination converged, final stess value: ', strs)
  else 
    txt <- paste0('Ordination did not converge, final stess value: ', strs)
  
  return(txt)

})
```

```{r ord1plt, out.width = "100%"}
renderPlot({
  
  # input
  clstcut <- as.numeric(input$clstcut)
  clstgrps <- clstgrps()
  
  # use watersheds as groups if clstcut is one, otherwise cluster groups
  if(clstcut == 1){
    
    grpin <- as.character(wshdord1())
    
    # subset colors by actual wsheds in ord
    colsub <- colgrp %>% 
      filter(Watershed %in% unique(wshdord1())) %>% 
      pull(cols)
    
  } else {
    
    grpin <- as.character(clstgrps)
    colsub <-  mapviewGetOption("vector.palette")(clstcut)
    
  }

  pord1 <- ggord(ord1(), grp_in = grpin, vec_ext = 1, col = colsub, 
        alpha = 1, alpha_el = 0.2, size = 4, txt = 5, arrow = 0.5, repel = F, coord_fix = F) + 
    theme_bw(base_family = 'serif', base_size = 16) +
    theme(legend.position = 'top')
  
  return(pord1)
  
}, width = 550, height = 600)
```

## Dissimilarity matrix

Pairwise comparisons of site dissimilarities based on species presence/absence.  Comparisons in red are more dissimilar, whereas those in blue are more similar.

```{r}
column(12, 
  column(4, NULL),
  # select matrix order type
  column(width = 4,
    selectInput('byclst', 'Order by?', selected = 'Clusters', choices = c('Watershed', 'Clusters'))
  ),
  column(4, NULL)
)
```

```{r dis1plt, out.width = "100%"}
renderPlot({
  
  # inputs
  byclst <- input$byclst
  dis1 <- dis1()
  clst <- clst()
  clstgrps <- clstgrps()

  # prep distance data to plot
  # long format of dist matrix, joind with wshed
  toplo <- dis1 %>% 
    as.matrix %>% 
    as.data.frame %>%
    rownames_to_column('SiteID') %>% 
    gather('SiteID2', 'dist', -SiteID) %>% 
    left_join(siteshd, by = 'SiteID') %>% 
    arrange(Watershed, dist) %>% 
    mutate(
      Watershed = factor(Watershed, levels = unique(Watershed)), 
      dist = ifelse(SiteID == SiteID2, NA, dist)
      )
  
  # get site order levels based on clustering
  if(byclst == 'Clusters'){
    
    sitfc <- clst$labels[clst$order]
    toplo <- toplo %>% 
      mutate(
        SiteID = factor(SiteID, levels = sitfc), 
        SiteID2 = factor(SiteID2, levels = sitfc)
      )
    
  } 
  
  # plot
  p <- ggplot(toplo) + 
    geom_tile(aes(x = SiteID, y = SiteID2, fill = dist), colour = 'black') +
    scale_x_discrete('', expand = c(0, 0)) + 
    scale_y_discrete('', expand = c(0, 0)) +
    scale_fill_gradient2('Dissimilarity between sites\nby species p/a', low = 'lightblue', mid = 'white', high = 'tomato1', midpoint = 0.5) +
    guides(fill = guide_colourbar(barheight = 0.5, barwidth = 10, label.theme = element_text(size = 11, angle = 0))) + 
    pbase
  
  # add vline/hline
  if(byclst == 'Watershed'){
      
    # index values of watershed divisions in plot
    brks <- toplo %>% 
      spread(SiteID2, dist) %>% 
      pull(Watershed) %>% 
      duplicated %>% 
      `!` %>% 
      which %>% 
      `-` (0.5) %>% 
      .[-1]
  
  } else {
    
    # index values of watershed divisions in plot
    brks <- clstgrps %>% 
      .[clst$order] %>% 
      duplicated %>% 
      `!` %>% 
      which %>% 
      `-` (0.5) %>% 
      .[-1]
    
  }
    
  # add watershed or cluster breaks to the plot
  p <- p + 
    geom_vline(xintercept = brks, size = 1.5) +
    geom_hline(yintercept = brks, size = 1.5)
  
  return(p)
  
}, width = 650, height = 700)
```

## Dendrogram

Cluster dendrogram colored by selected number of groups.

```{r clstpl}
# cluster analysis
clst <- reactive({
  
  # inputs
  clsttyp <- 'average' #input$clsttyp
  
  out <- dis1() %>% 
    hclust(method = clsttyp)
  
  return(out)
  
})

# get cluster groups
clstgrps <- reactive({
  
  # input
  clst <- clst()
  clstcut <- input$clstcut

  # get cut groups
  grps <- cutree(clst, k = clstcut)
  
  return(grps)
  
})
```

```{r dend}
# cluster dendrogram
renderPlot({

  # input
  clst <- clst()
  clstcut <- input$clstcut
  clstgrps <- clstgrps()

  # get order
  clstord <- order.hclust(clst) %>% 
    clstgrps[.] %>% 
    unique

  # get colors, correct by order
  cols <- mapviewGetOption("vector.palette")(as.numeric(clstcut)) %>% 
    .[clstord]
  
  p1 <- clst %>% 
    as.dendrogram %>% 
    set("branches_k_color", k = clstcut, value = cols) %>%
    set("labels_colors", k = clstcut, value = cols) %>%
    set("labels_cex", 0.8) 
  
  plot(p1)
  
})
```

## Cluster map

Sites are colored by the selected number of cluster groups (see dendrogram tab).

```{r mapclstprp}
# subset year and species,  map
mapclst <- reactive({
  
  # inputs
  clstgrps <- clstgrps()
  
  # clstgrps dataframe
  clstgrps <- clstgrps %>% 
    data.frame(clstgrps = .) %>% 
    rownames_to_column('SiteID')
  
  toplo <- spdat %>%
    dplyr::select(SiteID) %>% 
    filter(!duplicated(SiteID)) %>% 
    mutate(SiteID = as.character(SiteID)) %>% 
    inner_join(clstgrps, by = 'SiteID')

  out <- mapview(stream, label = stream$STREAM_NM, homebutton = F, legend = F) +
    mapview(toplo, zcol = 'clstgrps', legend = F, homebutton = F)

  return(out)
  
})
```

```{r mapclst}
renderLeaflet({mapclst()@map})
```

## Dominant species by group

These plots show the proportion of sites in a group where a species was observed.  This provides a measure of relative abundance based solely on presence/absence data. 

```{r}
renderPlot({
  
  # inputs
  clstgrps <- clstgrps()
  toord <- mltdatyr1() %>% 
    dplyr::select(-Watershed) 
  
  # toplo
  toplo <- toord %>% 
    mutate(clstgrps = clstgrps) %>% 
    gather('spp', 'pa', -clstgrps) %>% 
    group_by(clstgrps, spp) %>% 
    summarise(
      cnts = sum(pa)
      ) %>% 
    mutate(
      pers = 100 * cnts/sum(cnts)
    ) %>% 
    ungroup %>% 
    filter(pers > 0) %>% 
    unite('clstsp', clstgrps, spp, remove = F) %>% 
    arrange(clstgrps, pers) %>% 
    group_by(clstgrps) %>% 
    mutate(
      clstsp = factor(clstsp, levels = clstsp),
      perscum = cumsum(pers)
    )
      
  # plot
  p <- ggplot(toplo, aes(x = factor(clstgrps), fill = pers, group = factor(clstgrps))) +
    geom_bar(stat = 'identity', aes(y = pers), width = 0.25, colour = 'black') +
    geom_text_repel(aes(label = spp, y = perscum), angle = 0, nudge_x = 0.25, hjust = 0, direction = 'y') +
    scale_x_discrete('Group') + 
    scale_y_continuous('Cumulative occupied sites (%)') +
    scale_fill_distiller('%', palette = 'Spectral') +
    coord_flip() +
    theme_minimal(base_size = 18, base_family = 'serif') +
    theme(
      panel.grid.minor = element_blank(), 
      panel.grid.major = element_blank()
    )

  return(p)
  
}, height = 500, width = 900)
```