---
title: "Community analysis"
runtime: shiny
output: 
  html_document:
    toc: true
    toc_float: 
       collapsed: false
---

```{r setupcomman, include = FALSE}
# globals
library(knitr)
opts_chunk$set(echo = F, message = F, warning = F)

library(tidyverse)
library(sf)
library(mapview)
library(plotly)
library(shiny)
library(scales)
library(leaflet)
library(stargazer)
library(EnvStats)
library(kableExtra)

prj <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"

source('R/funcs.R')

data(habitat)
data(fishdat)
data(stream)

# prep species data, long format 
spdat <- fishdat %>% 
  select(Year, SiteID, matches('^Sp_')) %>% 
  gather('species', 'pa', -Year, -SiteID, -geometry) %>% 
  mutate(
    species = gsub('^Sp_', '', species)
  )
```

<br>

## Overview maps

This map shows species presence/absence in the four watersheds.  Sites that were sampled in a given year are shown on the map with small points showing a species was not found and large points showing a species was found.

```{r sppinput}
selectInput('yr', 'Select a year:', sort(unique(spdat$Year)))
selectInput('sp', 'Select a species:', sort(unique(spdat$species)))
```

```{r prepmapsp}
# subset year and species,  map
mapspyr <- reactive({
  
  yr <- input$yr
  sp <- input$sp
  
  toplo <- spdat %>% 
    filter(Year == yr & species == sp) %>% 
    mutate(
      cexv = ifelse(pa == 1, 7, 3), 
      pa = ifelse(pa == 1, 'present', 'absent')
      )
  
  out <- mapview(stream, label = stream$STREAM_NM) +
    mapview(toplo, zcol = 'pa', cex = toplo$cexv, legend = F)
   
  return(out)
})
```

```{r mapsp}
renderLeaflet({mapspyr()@map})
```

Plot of percent sites over time with a species, by watershed, n number of sites sampled in a year at the top.

Multivariate analysis of sites over time.

```{r test2}
plot(rnorm(100))
```

