---
title: "Community and habitat analysis"
runtime: shiny
output: 
  html_document:
    toc: true
    toc_float: 
       collapsed: false
    number_sections: true
---

```{r setupcommhaban, include = FALSE}
# globals
library(knitr)
opts_chunk$set(echo = F, message = F, warning = F)

library(tidyverse)
library(sf)
library(mapview)
library(shiny)
library(scales)
library(leaflet)
library(ggord)
library(vegan)
library(ggdendro)
library(dendextend)
library(shinyWidgets)
library(ggrepel)

source('R/funcs.R')

data(habitat)
data(fishdat)
data(stream)

prj <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"

# prep species data for multivariate analysis, long format 
spdat <- fishdat %>% 
  dplyr::select(Year, Watershed, SiteID, matches('^Sp_')) %>% 
  gather('species', 'pa', -Year, -SiteID, -Watershed, -geometry) %>% 
  mutate(
    species = gsub('^Sp_', '', species)
  ) %>% 
  filter(!is.na(Year)) %>% 
  group_by(Year, species) %>% 
  filter(!duplicated(SiteID)) %>% 
  ungroup() %>%
  unique

# prep species data for multivariate analysis
# same as spdat, but no geometry
mltspdat <- spdat
st_geometry(mltspdat) <- NULL

# prep habitat data for multivariate analysis
# same as habitat, but no geometry
mlthbdat <- habitat
st_geometry(mlthbdat) <- NULL

# site, wshed lookup
siteshd <- fishdat
st_geometry(siteshd) <- NULL
siteshd <- siteshd %>% 
  dplyr::select(SiteID, Watershed) %>% 
  unique %>% 
  mutate_if(is.factor, as.character)

# ggplot base 
pbase <- theme_bw(base_family = 'serif') +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 11), 
    axis.text.y = element_text(size = 11),
    legend.position = 'top',
    legend.direction = 'horizontal',
    # plot.margin = unit(c(4,4,0,0), "lines"),
    strip.background = element_blank(), 
    strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5), 
    panel.background = element_rect(fill = 'black')
  ) 

# habitat variables to select, master obj
habvrs <- list(
  'Canopy cover (%)' = 'StnCan',
  'Deciduous canopy cover (%)' = 'StnDecid', 
  'Average depth (ft)' = 'StnDpthAvg',
  'Maximum depth (ft)' = 'StnDpthMax',
  'Embeddedness (%)' = 'StnEmbed',
  'Escape cover (ratio)' = 'StnEsCov',
  'Fines (%)' = 'StnFines',
  'Station length (ft)' = 'StnLgth',
  'Station width (ft)' = 'StnWdth'
  )

# map colors
cols <- mapviewGetOption("vector.palette")(4)

# map colors for wsheds
colgrp <- data.frame(
  Watershed = c('APT', 'PAJ', 'SLR', 'SOQ'), 
  cols = cols, 
  stringsAsFactors = F
)
```

This section can be used to evaluate community changes over time and how these changes may relate to measured habitat variables.  The setup is similar as the analysis on the [community](comman.Rmd) page with some important differences.    

1) The species presence/absence (community) data are clustered as before, with an option to select the number of groups.
1) The comparison of community and habitat data is done using constrained ordination analysis that jointly considers relationships among species in the community data and how those relationships compare to patterns in the habitat data.  
1) The results from the last step can be compared to the groupings in the first step to identify if:
     * The groupings are still relevant when considering relationships with habitat,
     * If relevant, which habitat variables are important for species groupings?
1) A necessary limitation of the analysis is that sites must match between the community and habitat data.  This places additional restrictions on the data.

Some questions to consider when evaluating the data are:

1) Do the results differ when evaluating the results between years?
1) Is there an effect of adding or removing habitat variables from the analysis?
1) Can the results be explained by community or habitat differences between the watersheds?
1) If not, are there other natural groupings in the data that can describe community differences?

__Note that the habitat variables used in this analysis were those collected at the fish sampling sites.__

```{r mltsetup}
column(12, 
       
       column(2, NULL),
  
       # select a year
       column(width = 4, 
         selectInput('yrmlt', 'Select a year for comparison:', sort(unique(mlthbdat$Year)), selected = 2018)
       ),

       # select cluster cuts
       column(width = 4, 
         selectInput('clstcut', 'Select cluster groups:', selected = 2, choices = c(1:10))
       ),
      
       column(2, NULL)
       
)
```

```{r ccaord}
# species and habitat data for ordination 
# returns two element list with names toordsp, toordhb
# sites (rows) are matched between the species and habitat data
toord <- reactive({
  
  # inputs
  yrmlt <- input$yrmlt
  remrr <- 1 #input$remrr
  habvr <- input$habvr
  habty <- input$habty

  # create species pa mat after filtering species at less than remrr sites
  toordsp <- mltspdat %>% 
    filter(Year == yrmlt) %>% 
    dplyr::select(SiteID, Watershed, species, pa) %>% 
    group_by(species) %>% 
    mutate(totsp = sum(pa)) %>% 
    filter(totsp > remrr) %>% 
    dplyr::select(-totsp) %>% 
    spread(species, pa) %>% 
    data.frame(stringsAsFactors = F)
  rownames(toordsp) <- NULL
  toordsp <- toordsp %>% 
    column_to_rownames('SiteID')

  # remove sites with nothing found after rare spp removed
  torm <- rowSums(toordsp[, !names(toordsp) %in% c('Watershed')])
  toordsp <- toordsp[torm > 0, ]

  # create habitat data as input for ord
  # data are centered and scaled
  toordhb <- mlthbdat %>% 
    filter(Year == yrmlt) %>% 
    filter(habvar %in% habvr) %>% 
    filter(HabType %in% habty) %>% 
    dplyr::select(SiteID, Watershed, habvar, habval) %>% 
    spread(habvar, habval) %>% 
    data.frame(stringsAsFactors = F) %>% 
    mutate_if(is.numeric, base::scale) %>% 
    na.omit # removes rows where data are missing
  rownames(toordhb) <- NULL
  toordhb <- toordhb %>% 
    column_to_rownames('SiteID')
  
  # find matching sites and subset
  stmt <- dplyr::intersect(rownames(toordhb), rownames(toordsp)) %>% sort
  toordsp <- toordsp[stmt, ]
  toordhb <- toordhb[stmt, ]

  # output
  out <- list(toordsp = toordsp, toordhb = toordhb)  
  return(out)
  
})

# wshed groups for ggord
wshdord <- reactive({

  toord()[['toordsp']] %>% pull(Watershed)
  
})

# dissimilarity matrix
dis <- reactive({
  
  # input
  distyp <- 'jaccard' #input$distyp

  toordsp <- toord()[['toordsp']] %>% 
    dplyr::select(-Watershed) 
  
  # metaMDS
  dis <- toordsp %>% 
    vegdist(method = distyp)
  
  return(dis)
  
})

# ordination
ccaord <- reactive({
  
  # input
  toordsp <- toord()[['toordsp']]
  toordhb <- toord()[['toordhb']]
  # remove watershed column
  toordsp <- toordsp %>% 
    dplyr::select(-Watershed)
  toordhb <- toordhb %>% 
    dplyr::select(-Watershed)
  
  # frmin <- names(toordhb) %>% 
  #   paste(collapse = ' + ') %>% 
  #   paste('toordsp', ., sep = ' ~ ') %>% 
  #   as.formula
  
  # remove Stn prefix from habitat names
  names(toordhb) <- gsub('^Stn', '', names(toordhb))

  # metaMDS
  ccaord <- cca(toordsp, toordhb)
  
  return(ccaord)
  
})
```

# Species clustering results {.tabset}

## Dissimilarity matrix

Pairwise comparisons of site dissimilarities based on species presence/absence.  Comparisons in red are more dissimilar, whereas those in blue are more similar.

```{r}
column(12, 
  column(4, NULL),
  # select matrix order type
  column(width = 4,
    selectInput('byclst', 'Order by?', selected = 'Clusters', choices = c('Watershed', 'Clusters'))
  ),
  column(4, NULL)
)
```

```{r displt, out.width = "100%"}
renderPlot({
  
  # inputs
  byclst <- input$byclst
  dis <- dis()
  clst <- clst()
  clstgrps <- clstgrps()

  # prep distance data to plot
  # long format of dist matrix, joind with wshed
  toplo <- dis %>% 
    as.matrix %>% 
    as.data.frame %>%
    rownames_to_column('SiteID') %>% 
    gather('SiteID2', 'dist', -SiteID) %>% 
    left_join(siteshd, by = 'SiteID') %>% 
    arrange(Watershed, dist) %>% 
    mutate(
      Watershed = factor(Watershed, levels = unique(Watershed)), 
      dist = ifelse(SiteID == SiteID2, NA, dist)
      )
  
  # get site order levels based on clustering
  if(byclst == 'Clusters'){
    
    sitfc <- clst$labels[clst$order]
    toplo <- toplo %>% 
      mutate(
        SiteID = factor(SiteID, levels = sitfc), 
        SiteID2 = factor(SiteID2, levels = sitfc)
      )
    
  } 
  
  # plot
  p <- ggplot(toplo) + 
    geom_tile(aes(x = SiteID, y = SiteID2, fill = dist), colour = 'black') +
    scale_x_discrete('', expand = c(0, 0)) + 
    scale_y_discrete('', expand = c(0, 0)) +
    scale_fill_gradient2('Dissimilarity between sites\nby species p/a', low = 'lightblue', mid = 'white', high = 'tomato1', midpoint = 0.5) +
    guides(fill = guide_colourbar(barheight = 0.5, barwidth = 10, label.theme = element_text(size = 12, angle = 0))) + 
    pbase
  
  # add vline/hline
  if(byclst == 'Watershed'){
      
    # index values of watershed divisions in plot
    brks <- toplo %>% 
      spread(SiteID2, dist) %>% 
      pull(Watershed) %>% 
      duplicated %>% 
      `!` %>% 
      which %>% 
      `-` (0.5) %>% 
      .[-1]
  
  } else {
    
    # index values of watershed divisions in plot
    brks <- clstgrps %>% 
      .[clst$order] %>% 
      duplicated %>% 
      `!` %>% 
      which %>% 
      `-` (0.5) %>% 
      .[-1]
    
  }
    
  # add watershed or cluster breaks to the plot
  p <- p + 
    geom_vline(xintercept = brks, size = 1.5) +
    geom_hline(yintercept = brks, size = 1.5)
  
  return(p)
  
}, width = 650, height = 700)
```

## Dendrogram

Cluster dendrogram colored by selected number of groups.

```{r clstpl}
# cluster analysis
clst <- reactive({
  
  # inputs
  clsttyp <- 'average'
  
  out <- dis() %>% 
    hclust(method = clsttyp)
  
  return(out)
  
})

# get cluster groups
clstgrps <- reactive({
  
  # input
  clst <- clst()
  clstcut <- input$clstcut

  # get cut groups
  grps <- cutree(clst, k = clstcut)
  
  return(grps)
  
})
```

```{r dend}
# cluster dendrogram
renderPlot({
  
  # input
  clst <- clst()
  clstcut <- input$clstcut
  clstgrps <- clstgrps()

  # get order
  clstord <- order.hclust(clst) %>% 
    clstgrps[.] %>% 
    unique

  # get colors, correct by order
  cols <- mapviewGetOption("vector.palette")(as.numeric(clstcut)) %>% 
    .[clstord]
  
  p1 <- clst %>% 
    as.dendrogram %>% 
    set("branches_k_color", k = clstcut, value = cols) %>%
    set("labels_colors", k = clstcut, value = cols) %>%
    set("labels_cex", 0.8) 
  
  plot(p1)
  
})
```

## Cluster map

Sites are colored by the selected number of cluster groups (see dendrogram tab).

```{r mapclstprp}
# subset year and species,  map
mapclst <- reactive({
  
  # inputs
  clstgrps <- clstgrps()
  
  # clstgrps dataframe
  clstgrps <- clstgrps %>% 
    data.frame(clstgrps = .) %>% 
    rownames_to_column('SiteID')
  
  toplo <- spdat %>%
    dplyr::select(SiteID) %>% 
    unique %>% 
    mutate(SiteID = as.character(SiteID)) %>% 
    inner_join(clstgrps, by = 'SiteID')

  out <- mapview(stream, label = stream$STREAM_NM, homebutton = F) +
    mapview(toplo, zcol = 'clstgrps', legend = F, homebutton = F)

  return(out)
  
})
```

```{r mapclst}
renderLeaflet({mapclst()@map})
```

# Constrained ordination {.tabset}

```{r}

column(12, 
       
      column(2, NULL),
       
      # habitat type to evaluate
      column(width = 4, 
             selectInput("habty", label = 'Choose habitat type:', choices = c('run', 'riffle', 'pool'), selected  = 'riffle')
      ),

      # habitat variables to consider
      column(width = 4, 
        pickerInput("habvr", label = 'Choose habitat variables:', choices = habvrs, selected = habvrs,
                                                      multiple = TRUE
                                          )
        ),
      
      column(2, NULL)
      
)
```

## Triplot

The ordination below shows results for constrained correspondence analysis, which is similar to the ordination on the [community](comman.Rmd) page with an additional multivariate constraint for the habitat variables.  Species associations are grouped relative to the habitat variables and the groups can be interpreted relative to both.  

```{r selax}

column(12, 
       
    column(2, NULL),
       
    column(4, 
      renderUI({
        
        # inputs
        ccaord <- ccaord()
        
        # possible axes to select
        axs <- seq_along(colnames(ccaord$CCA$wa))
      
        selectInput('axsx', 'Select CCA axis for x-axis:', choices = axs, selected = 1)
        
      })
    ),

    column(4, 
      renderUI({
        
        # inputs
        ccaord <- ccaord()
        
        # possible axes to select
        axs <- seq_along(colnames(ccaord$CCA$wa))
      
        selectInput('axsy', 'Select CCA axis for y-axis:', choices = axs, selected = 2)
        
      })
    ),
    
    column(2, NULL)

)
```

```{r ordplt, out.width = "100%"}
renderPlot({

  req(input$axsx)
  
  # input
  ccaord <- ccaord()
  clstcut <- as.numeric(input$clstcut)
  clstgrps <- clstgrps()
  axsx <- input$axsx
  axsy <- input$axsy

  # use watersheds as groups if clstcut is one, otherwise cluster groups
  if(clstcut == 1){

    grpin <- as.character(wshdord())

    # subset colors by actual wsheds in ord
    colsub <- colgrp %>%
      filter(Watershed %in% unique(wshdord())) %>%
      pull(cols)

  } else {

    grpin <- as.character(clstgrps)
    colsub <-  mapviewGetOption("vector.palette")(clstcut)

  }

  pord <- try({ggord(ccaord, grp_in = grpin, axes = c(axsx, axsy), vec_ext = 4, col = colsub, ptslab = T, parse = T,
        alpha = 1, alpha_el = 0.2, size = 4, txt = 5, addsize = 4, arrow = 0.5, repel = F, coord_fix = F) +
    theme_bw(base_family = 'serif', base_size = 16) +
    theme(legend.position = 'top')})

  validate(
    need(!inherits(pord, 'try-error'), 'Insufficient data')
    )
    
  return(pord)

}, width = 700, height = 700)
```

## Dominant species by group

These plots show the proportion of sites in a group where a species was observed.  This provides a measure of relative abundance based solely on presence/absence data. 

```{r}
renderPlot({
  
  # inputs
  clstgrps <- clstgrps()
  toordsp <- toord()[['toordsp']]

  # toplo
  toplo <- toordsp %>% 
    dplyr::select(-Watershed) %>% 
    mutate(clstgrps = clstgrps) %>% 
    gather('spp', 'pa', -clstgrps) %>% 
    group_by(clstgrps, spp) %>% 
    summarise(
      cnts = sum(pa)
      ) %>% 
    mutate(
      pers = 100 * cnts/sum(cnts)
    ) %>% 
    ungroup %>% 
    filter(pers > 0) %>% 
    unite('clstsp', clstgrps, spp, remove = F) %>% 
    arrange(clstgrps, pers) %>% 
    group_by(clstgrps) %>% 
    mutate(
      clstsp = factor(clstsp, levels = clstsp),
      perscum = cumsum(pers)
    )
      
  # plot
  p <- ggplot(toplo, aes(x = factor(clstgrps), fill = pers, group = factor(clstgrps))) +
    geom_bar(stat = 'identity', aes(y = pers), width = 0.25, colour = 'black') +
    geom_text_repel(aes(label = spp, y = perscum), angle = 0, nudge_x = 0.25, hjust = 0, direction = 'y') +
    scale_x_discrete('Group') + 
    scale_y_continuous('Cumulative occupied sites (%)') +
    scale_fill_distiller('%', palette = 'Spectral') +
    coord_flip() +
    theme_minimal(base_size = 18, base_family = 'serif') +
    theme(
      panel.grid.minor = element_blank(), 
      panel.grid.major = element_blank()
    )

  return(p)
  
}, height = 500, width = 900)
```

## Relative habitat variables by group

These plots show the relative associations of habitat variables with the selected groups.  All habitat variables were scaled and centered to have zero mean and unit variance (standard deviation = 1) for ordination.  This allows for relative comparisons between the magnitudes of the values associated with each habitat variable, i.e., negative values are lower and positive values are higher.
```{r}
renderPlot({
  
  # inputs
  clstgrps <- clstgrps()
  toordhb <- toord()[['toordhb']]

  # toplo
  toplo <- toordhb %>%
    dplyr::select(-Watershed) %>%
    data.frame(clstgrps = clstgrps) %>% 
    gather('var', 'val', -clstgrps) %>%
    group_by(clstgrps, var) %>% 
    summarise(val = mean(val)) %>% 
    ungroup %>% 
    mutate(var = gsub('^Stn', '', var))
   
  # plot
  p <- ggplot(toplo, aes(x = factor(var), y = val, fill = val)) +
    geom_bar(stat = 'identity', width = 0.25, colour = 'black') + 
    facet_wrap(~clstgrps) +
    geom_hline(yintercept = 0) +
    # scale_y_continuous('Cumulative occupied sites (%)') +
    scale_fill_gradient2(low = 'tomato1', high = 'lightgreen', mid = 'white', midpoint = 0) +
    coord_flip() +
    theme_minimal(base_size = 18, base_family = 'serif') +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(), 
      strip.background = element_blank(), 
      axis.title.y = element_blank(), 
      legend.position = 'none'
    ) +
    scale_y_continuous('Relative value')

  return(p)
  
}, height = 500, width = 900)
```