---
title: "Habitat analysis - site"
runtime: shiny
output: 
  html_document:
    toc: true
    toc_float: 
       collapsed: false
    number_sections: true
---

```{r setuphaban, include = FALSE}
# globals
library(knitr)
opts_chunk$set(echo = F, message = F, warning = F)

library(tidyverse)
library(sf)
library(mapview)
library(shiny)
library(scales)
library(leaflet)
library(ggord)
library(vegan)
library(ggdendro)
library(dendextend)
library(shinyWidgets)
library(EnvStats)

source('R/funcs.R')

data(habitat)
data(stream)
data(trndhab_prep)

prj <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"

# habitat type colors
habtycol <- colorRampPalette(c('tomato1', 'lightgreen'))(3)
names(habtycol) <- c('run', 'riffle', 'pool')

# habitat variables to select, master obj
habvrs <- list(
  'Canopy cover (%)' = 'StnCan',
  'Deciduous canopy cover (%)' = 'StnDecid', 
  'Average depth (ft)' = 'StnDpthAvg',
  'Maximum depth (ft)' = 'StnDpthMax',
  'Embeddedness (%)' = 'StnEmbed',
  'Escape cover (ratio)' = 'StnEsCov',
  'Fines (%)' = 'StnFines',
  'Station length (ft)' = 'StnLgth',
  'Station width (ft)' = 'StnWdth'
  )
```

<br>

These analyses can be used to evaluate changes in habitat over time, by watershed, habitat type, and habitat measurements.  The habitat data in this section were collected at sites where fish sampling also occurred. The first two sections show general changes by individual years and by watershed.  The third section shows more detailed changes for individual sites for selected periods of time. 

Habitat measurements were taken separately in different habitat types at each site:

* __Run__
* __Riffle__
* __Pool__

The analyses are structured around these habitat types.  Some habitat types are designated as combinations of the above but these are in the minority and are not included in the analyses below.

```{r rctvs}
# filtered habitat data by type and variable
habrct <- reactive({
  
  # input
  habtyp <- input$habtyp
  habvr <- input$habvr

  out <- habitat %>% 
    filter(HabType %in% habtyp) %>% 
    filter(habvar %in% habvr)
  
  return(out)
  
})

# map showing selected habitat type, variable, and year
habmap <- reactive({

  # input
  yr <- input$yr
  habrct <- habrct()

  habrct <- habrct %>% 
    filter(Year %in% yr)
  
  # hover pt labels
  labs <- paste(habrct$SiteID, ': ', habrct$habval)
  
  out <- mapview(stream, label = stream$STREAM_NM, homebutton = F, legend = F) +
    mapview(habrct, zcol = 'Watershed', cex = 'habval', label = labs, legend = T, layer.name = 'Watershed', homebutton = F)
  
  return(out)
  
})
```

# Overview maps

This map shows the relative values for a selected habitat variable and year for the chosen habitat type. Observations are colored by watershed and points are sized by relative values of the selected variable.

```{r habinput}
column(12, 
  column(2, NULL),
  # select habitat variable to show
  column(width = 4,
        selectInput('habvr', 'Select a habitat variable:', choices = habvrs)
  ),
  # select year to show
  column(width = 4,
        selectInput('yr', 'Select a year:', selected = 2018, choices = sort(unique(habitat$Year)))
  ),
  column(2, NULL)
)
column(12, 
  column(2, NULL),
  # habitat type to evaluate
  column(width = 4,
      selectInput("habtyp", "Choose a habitat type:", choices = c('run', 'riffle', 'pool'))
  ),
  column(6, NULL)
)
```

```{r}
renderLeaflet({habmap()@map})
```

<br>

# Watershed averages over time

This plot shows measured habitat variables over time, averaged across all sites that were measured each year in a given watershed. Habitat types (run, riffle, pool) can be plotted individually or together.  The numbers inside the boxes indicate the number of sites that were averaged for each year, watershed combination.  Note that the number of sites that were visited in each watershed varied by year and the trends must be interpreted accordingly.  Overlaying the observations can aid the interpretation.

```{r tmsel}
column(width = 12,
       
  column(2, NULL),
       
  # habitat type to evaluate
  column(width = 4, 
  pickerInput("habtyp2", label = 'Choose a habitat type:', choices = c('run', 'riffle', 'pool'),
                                                selected = c('run'),
                                                multiple = TRUE
                                    )
  ),
  
  # select habitat variable to show
  column(width = 4,
        selectInput('habvr2', 'Select a habitat variable:', choices = habvrs)
  ),
  
  column(2, NULL)
  
)

column(12, 
       
  column(2, NULL),
       
  # year slider
  column(width = 4, 
    sliderInput("yrrng", label = 'Select year ranges to plot:',  
          min = 2001, max = 2018, 
          value = c(2001, 2018),
          sep = '', ticks = FALSE
        )
  ),
  
  # show actual obs in each wshed or not
  column(width = 4, 
    selectInput("shwobs", label = 'Overlay observed values?', choices = list('no' = F, 'yes' = T))
  ),
  
  column(2, NULL)
  
)
```

```{r habyragg}
# all values of a selected habitat variable over time, by watershed
habyrall <- reactive({
  
  # inputs
  habvr2 <- input$habvr2
  habtyp2 <- input$habtyp2
  yrrng <- input$yrrng

  out <- habitat
  st_geometry(out) <- NULL
  out <- out %>% 
    filter(habvar %in% habvr2) %>% 
    filter(Year >= yrrng[1] & Year <= yrrng[2]) %>% 
    filter(HabType %in% habtyp2) %>% 
    na.omit
  
  return(out)
  
})

# average values of a selected habitat variable over time, by watershed
habyragg <- reactive({
  
  # inputs
  habyrall <- habyrall()

  out <- habyrall %>% 
    group_by(Year, HabType, Watershed) %>% 
    summarise(
      n = n(), 
      avehab = mean(habval, na.rm = T)
    ) %>% 
    ungroup %>% 
    mutate_if(is.factor, as.character) %>% 
    complete(Year = full_seq(Year, 1), HabType, Watershed)
  
  return(out)
  
})
```

```{r prsplo, out.width = "100%"}
renderPlot({
  
  # inputs
  habyragg <- habyragg()
  habyrall <- habyrall()
  habvr2 <- input$habvr2
  shwobs <- input$shwobs
  
  # breaks on x-axis
  yrbrks <- range(habyragg$Year, na.rm = T)
  yrbrks <- seq(yrbrks[1], yrbrks[2])
  
  # plot mapping
  p <- ggplot(habyragg, aes(x = Year, group = HabType, fill = HabType, color = HabType)) 
  
  # add obs if shwobs
  if(shwobs){

     p <- p + 
       geom_jitter(data = habyrall, aes(y = habval), size = 1.5, width = 0.05, alpha = 0.8)
     
  } 
  
  # create rest of plot
  p <- p + 
    stat_smooth(aes(y = avehab), method = 'lm', se = F, linetype = 'dashed') +
    geom_label(aes(y = avehab, label = n), size = 5, colour = 'black') +
    facet_wrap(~Watershed, ncol = 2) +
    theme_bw(base_family = 'serif', base_size = 18) +
    theme(
      strip.background = element_blank(),
      axis.title.x = element_blank(),
      legend.position = 'top', 
      legend.title = element_blank(), 
      panel.grid.minor = element_blank(), 
      panel.grid.major = element_blank(), 
      axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1, size = 12)
    ) +
    scale_fill_manual(values = habtycol) +
    scale_colour_manual(values = habtycol) +
    scale_y_continuous(habvr2) + 
    scale_x_continuous(breaks = yrbrks)

  return(p)

}, width = 900, height = 700)

```

<br>

# Trends at individual sites {.tabset}

This section evaluates trends in selected habitat variables at individual sites, rather than aggregated by watershed as above.  Select the habitat type, habitat variable, and year range to evaluate the trends.  The map tab shows estimated trends across all sites in the selected years and the plot tab shows individual sites in more detail. 

```{r trndsel}

column(12, 
       
        column(2, NULL),
       
        # habitat type to evaluate
        column(width = 4, 
               selectInput("habtyp3", label = 'Choose a habitat type:', choices = c('run', 'riffle', 'pool'))
        ),

        # select habitat variable to show
        column(width = 4,
              selectInput('habvr3', 'Select a habitat variable:', choices = habvrs)
        ),
       
       column(2, NULL)
       
)

# year slider
column(width = 12, 
       
  column(2, NULL),
  
  column(4, 
         sliderInput("yrrng2", label = 'Select year ranges to plot:',  
          min = 2001, max = 2018, 
          value = c(2001, 2018),
          sep = '', ticks = FALSE
        )
  ), 
  
  column(6, NULL)
  
)
```

## Map, all sites

The points on the map show the value for tau (direction of trend) for a Kendall test of changes of the selected habitat variable within the selected years, green for increasing and red for decreasing.  Size of the point is the magnitude of the estimated change. Note that some sites may not have data spanning the full range of selected years.  Click an individual site on the map to see the available years within the selected range. 

```{r trndrcts}
# all trends for each site, selected years, habitat type, habitat variable, for map
mapdattrnd <- reactive({
  
  # input
  habtyp3 <- input$habtyp3
  habvr3 <- input$habvr3
  yrrng2 <- input$yrrng2

  toplo <- trndhab_prep %>% 
    filter(habvar %in% habvr3) %>% 
    mutate(
      trnd = purrr::map(data, function(x){

        xsub <- x %>% 
          filter(Year >= yrrng2[1] & Year <= yrrng2[2]) %>%
          filter(!is.na(habval)) %>% 
          mutate(
            aveval = mean(habval, na.rm = T), 
            habval = habval - aveval
            ) %>% 
          dplyr::select(-aveval)
        
        # get unique years tested
        yrs <- unique(xsub$Year) %>% 
          paste(collapse = ', ')
  
        # kendall sum
        knout <- suppressWarnings(try({kendallTrendTest(habval ~ Year, xsub)}, silent = T))
        if(inherits(knout, 'try-error')) return(NA)
        kpval <- knout$p.value
        kest <- knout$estimate
        out <- c(kpval, kest) %>% 
          as.list %>% 
          data.frame %>% 
          mutate(yrs = yrs)
           
        return(out)
  
        })
      ) %>% 
      dplyr::select(-data) %>% 
      filter(map_lgl(trnd, ~ !anyNA(.x))) %>% # this!
      unnest %>% 
      dplyr::select(-intercept) %>% 
      mutate(
        z = p_ast(z), 
        trend = sign(tau),
        trend = factor(trend, levels = c(-1, 1), labels =c('dec', 'inc'))
        ) %>% 
      mutate_if(is.numeric, round, 2) %>% 
      rename(pval = z) %>% 
      st_as_sf(coords = c("X", "Y"), crs =prj)
    
    return(toplo)
  
    })

# create map from selected, trended data
mapouthab <- reactive({

  # input
  mapdattrnd <- mapdattrnd()
  habtyp3 <- input$habtyp3
  
  # filter by selected habitat type
  mapdattrnd <- mapdattrnd %>% 
    filter(HabType %in% habtyp3)
  
  # get colors
  cols <- mapdattrnd %>%
    mutate(
      cols = factor(trend, levels = c('dec', 'inc'), labels = c('tomato1', 'lightgreen')),
      cols = as.character(cols)
    ) %>%
    pull(cols)
  
  # size values
  cexv <- mapdattrnd %>% 
    pull(tau) %>% 
    abs %>% 
    scales::rescale(to = c(2, 15))

  # hover pt labels
  labs <- paste(mapdattrnd$SiteID, ': ', mapdattrnd$trend, ', tau = ', mapdattrnd$tau, ', p = ', mapdattrnd$pval)
  
  out <- mapview(stream, label = stream$STREAM_NM, homebutton = F, legend = F) +
    mapview(mapdattrnd, zcol = 'trend', cex = cexv, label = labs, col.regions = cols, legend = F, homebutton = F)

  return(out)

  })

```
```{r habtrndmap}
# render the habitat trend map  
renderLeaflet({mapouthab()@map})
```

## Plot, one site

Trends for the measured habitat variable at a selected site are shown below.  Deviations from the mean are shown, where the means are estimated separately for different habitat types. These plots can confirm the results shown in the map.  

```{r habtrndpltprp}

column(12,
  
  column(2, NULL),
  
  # dynamic ui that changes with year selections on map, differents sites depending on year combos
  column(4, 
         renderUI({
   
          # inputs
          mapdattrnd <- mapdattrnd()
          habtyp3 <- input$habtyp3
          
          sts <- mapdattrnd() %>% 
            filter(HabType %in% habtyp3) %>% 
            pull(SiteID) %>% 
            as.character %>% 
            unique %>% 
            sort
          
          selectInput("st", "Select station to view trends:", sts)
        
        })
  ),
  
  # select which values to show on y-axis, deviations or actual
  column(4, 
  
          selectInput("yvr", "Select y-axis type:", c('deviation from mean', 'actual'))
  
  ),
  
  column(2, NULL)
  
)

# habitat trends to plot for an individual site, separate for habitat types
trndathab <- reactive({

  # input
  yrsel <- input$yearsst
  stsel <- input$st
  habvr3 <- input$habvr3
  yrrng2 <- input$yrrng2

  req(!is.null(stsel))
  
  toplo <- trndhab_prep %>% 
    filter(habvar %in% habvr3) %>% 
    filter(SiteID %in% stsel) %>% 
    mutate(
      habvaldf = purrr::map(data, function(x){

        out <- x %>% 
          filter(Year >= yrrng2[1] & Year <= yrrng2[2]) %>%
          filter(!is.na(habval)) %>% 
          mutate(
            aveval = mean(habval, na.rm = T), 
            habvaldf = habval - aveval
            ) %>% 
          dplyr::select(-aveval)
        
        return(out)
        
      })
    ) %>% 
    dplyr::select(-data) %>% 
    unnest
  
  return(toplo)

  })
```

```{r sitetrndplt, fig.height = 4, fig.width = 8, out.width = "100%"}
# plot of year trends at a site
renderPlot({
  
  toplo <- trndathab()
  mapdattrnd <- mapdattrnd()
  stsel <- input$st
  habvr3 <- input$habvr3
  yvr <- input$yvr
 
  # get trends, join with toplo
  trnds <- mapdattrnd %>%
    filter(SiteID %in% stsel) %>% 
    select(HabType, pval) %>% 
    mutate(
      pval = paste('trend', pval)
      )
  st_geometry(trnds) <- NULL
  toplo <- left_join(toplo, trnds, by = 'HabType') %>% 
    unite('HabType', HabType, pval, sep = ', ')
  
  req(!is.null(stsel))

  if(yvr == 'actual'){
    
    # midpoint for fill scaling    
    midpt <- mean(toplo$habval, na.rm = T)
    
    p <- ggplot(toplo, aes(x = Year, y = habval, fill = habval)) + 
      geom_bar(stat = 'identity', colour = 'grey') + 
      scale_fill_gradient2('', low = 'tomato1', mid = 'white', high = 'lightgreen', midpoint = midpt) +
      geom_smooth(method = 'lm', se = F, linetype = 'dashed', color = 'black') 
    
  } else {
    
    p <- ggplot(toplo, aes(x = Year, y = habvaldf, fill = habvaldf)) + 
      geom_bar(stat = 'identity', colour = 'grey') + 
      scale_fill_gradient2('', low = 'tomato1', mid = 'white', high = 'lightgreen', midpoint = 0) +
      geom_smooth(method = 'lm', se = F, linetype = 'dashed', color = 'black') + 
      geom_hline(yintercept = 0) 
        
  }
  
  # for axis format
  yrbrks <- range(toplo$Year, na.rm = T)
  yrbrks <- seq(yrbrks[1], yrbrks[2])
  
  p <- p +  
    facet_wrap(~ HabType, ncol = 3) + 
    theme_bw(base_family = 'serif', base_size = 18) +
    theme(
      strip.background = element_blank(),
      panel.grid = element_blank(), 
      axis.title.x = element_blank(), 
      axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1, size = 12)
    ) + 
    scale_y_continuous(habvr3) + 
    scale_x_continuous(breaks = yrbrks)

  return(p)
  
}, width = 900, height = 300)

```

<br>