---
title: "Evaluation of flow trends at Pajaro"
runtime: shiny
output: 
  html_document:
    toc: true
    toc_float: 
      collapsed: false
    number_sections: true
---

```{r setup, include = FALSE}
# globals
library(knitr)
opts_chunk$set(echo = F, message = F, warning = F)

library(tidyverse)
library(lubridate)
library(sf)
library(mapview)
library(plotly)
library(shiny)
library(scales)
library(leaflet)
library(stargazer)
library(EnvStats)
library(kableExtra)
library(patchwork)
library(shinyWidgets)

prj <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"

source('R/funcs.R')

data(fishdat)
data(fishmtch)
data(floest)

pbase <- theme_bw(base_family = 'serif', base_size = 14) +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1, size = 12), 
    axis.text.y = element_text(size = 12),
    legend.position = 'top',
    legend.direction = 'horizontal',
    # plot.margin = unit(c(4,4,0,0), "lines"),
    strip.background = element_blank(), 
    strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
    legend.text = element_text(size = 12)
  ) 

# fishmtch for pajaro
fishmtch <- fishmtch %>% 
  filter(Watershed %in% 'PAJ') %>% 
  dplyr::select(-Watershed)

# flow only
pajflo <- floest %>% 
  filter(Site %in% fishmtch$SiteFlow) %>% 
  filter(date > ymd('1997-01-01')) %>% 
  mutate(
    Year = year(date),
    mo = as.character(month(date, abbr = TRUE, label = T))
  ) %>% 
  unite('mosite', mo, Site, sep = ' - ', remove = F) 

# fish only
pajfsh <- fishdat %>% 
  filter(grepl('PAJ', Watershed)) %>% 
  dplyr::select(Year, SiteID, SampleDate, Dens_S1, Dens_S2) %>% 
  gather('Size class', 'density', Dens_S1, Dens_S2) %>% 
  filter(!is.na(density)) 

# flow and fish aggregated, combined
# standardize dates to september when fish sampling occurs to match with flow
pajbth <- pajfsh %>% 
  st_set_geometry(NULL) %>% 
  filter(month(SampleDate) >= 8) %>% 
  filter(SampleDate > ymd('1997-01-01')) %>% 
  mutate(
    mo = 9,
    dy = 15,
    SiteID = as.character(SiteID)
    ) %>% 
  unite('date', Year, mo, dy, sep = '-', remove = F) %>% 
  mutate(date = ymd(date)) %>% 
  inner_join(fishmtch, by = c('SiteID', 'mo')) %>% 
  mutate(mo = 'Sep') %>% 
  unite('mosite', mo, SiteFlow, sep = ' - ', remove = F) %>% 
  inner_join(pajflo, by = c('mosite', 'date')) %>% 
  dplyr::select(Group, SiteID, Site, date, `Size class`, density, flo) %>% 
  mutate(Year = year(date))
```

```{r rctvs}
fshyrs <- reactive({
  
  # input
  yrssel <- input$yrssel
  sizsel <- input$sizsel

  out <- pajfsh %>% 
    filter(Year >= yrssel[1] & Year <= yrssel[2]) %>% 
    filter(`Size class` %in% sizsel) %>% 
    group_by(SiteID) %>% 
    mutate(
      avedens = mean(density, na.rm = T), 
      density_df = density - avedens
      ) %>% 
    group_by(SiteID) %>% 
    filter(n() >= 3) %>% # kendall test will break if less than 3 obs
    nest %>% 
    mutate(
      tnd = purrr::map(data, function(x){

        knout <- kendallTrendTest(density_df ~ Year, x)
        outest <- round(knout$estimate, 2)
        outpval <- p_ast(knout$p.value) %>% 
          factor(levels = c('ns', '*', '**'))
        out <- tibble(tau = outest['tau'], slope = outest['slope'], pval = outpval)
        
        return(out)

      })
    )
  
  return(out)
    
})

floyrs <- reactive({
  
  # input
  yrssel <- input$yrssel

  out <- pajflo %>% 
    filter(Year >= yrssel[1] & Year <= yrssel[2]) %>% 
    group_by(mosite) %>% 
    mutate(
      aveflo = mean(flo, na.rm = T), 
      flo_df = flo - aveflo
      ) %>% 
    group_by(mosite) %>% 
    nest %>% 
    mutate(
      tnd = purrr::map(data, function(x){

        knout <- kendallTrendTest(flo_df ~ Year, x)
        outest <- round(knout$estimate, 2)
        outpval <- p_ast(knout$p.value) %>% 
          factor(levels = c('ns', '*', '**'))
        out <- tibble(tau = outest['tau'], slope = outest['slope'], pval = outpval)
        
        return(out)

      })
    )
  
  return(out)
  
})

fshind <- reactive({
  
  # input
  sitfsh <- input$sitfsh
  fshyrs <- fshyrs()
  
  out <- fshyrs %>% 
    filter(SiteID %in% sitfsh) %>% 
    dplyr::select(-tnd) %>% 
    unnest
  
  return(out)
  
})

floind<- reactive({
  
  # input
  sitflo <- input$sitflo
  floyrs <- floyrs()

  out <- floyrs %>% 
    filter(mosite %in% sitflo) %>% 
    dplyr::select(-tnd) %>% 
    unnest

  return(out)  
  
})

flotxt <- reactive({
  
  # input 
  floyrs <- floyrs()
  yrssel <- input$yrssel
  
  sigs <- floyrs %>% 
    dplyr::select(-data) %>% 
    unnest %>% 
    filter(!pval %in% 'ns')

  if(nrow(sigs) == 0)
    return(paste0('Zero flow records had significant trends from ', yrssel[1], ' to ', yrssel[2], '.'))
  
  if(nrow(sigs) == 1)
    out <- 'One flow record'

  if(nrow(sigs) > 1)
    out <- english::as.english(nrow(sigs)) %>% paste(., 'flow records')
  
  out <- out %>% 
    paste0(., ' had significant trends from ', yrssel[1], ' to ', yrssel[2] , ', ', english::as.english(sum(sigs$tau < 0)), ' decreasing and ', english::as.english(sum(sigs$tau >= 0)), ' increasing.')
  
  out <- paste('Trends:', out)
  
  return(out)
  
})

fshtxt <- reactive({
  
  # input 
  fshyrs <- fshyrs()
  yrssel <- input$yrssel
  
  sigs <- fshyrs %>% 
    dplyr::select(-data) %>% 
    unnest %>% 
    filter(!pval %in% 'ns')

  if(nrow(sigs) == 0)
    return(paste0('Zero fish sites had significant trends from ', yrssel[1], ' to ', yrssel[2], '.'))
  
  if(nrow(sigs) == 1)
    out <- 'One fish site'

  if(nrow(sigs) > 1)
    out <- english::as.english(nrow(sigs)) %>% paste(., 'fish sites')
  
  out <- out %>% 
    paste0(., ' had significant trends from ', yrssel[1], ' to ', yrssel[2] , ', ', english::as.english(sum(sigs$tau < 0)), ' decreasing and ', english::as.english(sum(sigs$tau >= 0)), ' increasing.')
  
  out <- paste('Trends:', out)
  
  return(out)
  
})

pajbthyrs <- reactive({
  
  # input
  yrssel <- input$yrssel
  pajgrp <- input$pajgrp
  
  out <- pajbth %>% 
    filter(Year >= yrssel[1] & Year <= yrssel[2]) %>% 
    filter(Group %in% pajgrp)
  
  return(out)
  
})

pajtndmod <- reactive({
  
  # input
  pajbthyrs <- pajbthyrs()

  req(nrow(pajbthyrs) > 0)
  
  out <- pajbthyrs %>% 
    group_by(`Size class`) %>% 
    nest %>%
    mutate(
      mod = purrr::map(data, function(x){
      
        mod <- lm(log10(1 + density) ~ log10(1 + flo), x)
        
        return(mod)
        
      })
    ) %>% 
    pull(mod)
  
  return(out)
  
})
```

```{r}
column(12, 
column(2, NULL),
column(4, 
  sliderInput("yrssel", label = 'Select year ranges:',  
        min = 1997, max = 2018, 
        value = c(1997, 2018),
        sep = '', ticks = FALSE
      )
),
column(4,
       selectInput('sizsel', label = 'Select size class:', choices = c('Dens_S1', 'Dens_S2'))
       
       ), 
column(2, NULL)
)
```

This page can be used to compare steelhead density estimates with modelled flow in the Pajaro watershed.  The sections include an assessment of overall trends, summary trends, plots by site, and combined trends for fish density and flow.  The year slider can be used to choose the date range for all plots and trend analyses.  The size class selector shows which size class is evaluated for steelhead.

# Overall trends {.tabset}

These plots show individual time series for flow and fish at all selected sites. The left plot shows observed values and the right plot shows differences in the observed values from the station or site average within the selected years.  Green indicates above average and red indicated below average.  The left plots also include a grey horizontal line for the site average and a dotted line for the linear trend.

## Flow

```{r}
renderPlot({
  
  # input
  floyrs <- floyrs()
  
  toplo <- floyrs %>% 
    dplyr::select(-tnd) %>% 
    unnest

  pall <- ggplot(toplo, aes(x = date, y = flo, fill = flo_df)) + 
    geom_hline(aes(yintercept = aveflo), colour = 'grey') +
    geom_line() + 
    geom_point(pch = 21, colour = 'black', size = 3) + 
    geom_smooth(method = 'lm', se = F, linetype = 'dashed', colour = 'black') +
    facet_grid(mosite ~ .) +
    pbase +
    theme(
      axis.title.x = element_blank(),
      legend.position = 'none',   
      axis.text.y = element_text(size = 8)
      ) +
    scale_y_log10('log-Flow (cfs)') + 
    scale_fill_gradient2('Difference from average,\nby location', low = 'tomato1', mid = "white", high = 'lightgreen', midpoint = 0) +
    guides(fill = guide_colourbar(barheight = 0.5, barwidth = 5, label.theme = element_text(size = 10, angle = 0)))
  
  toplorev <- toplo %>%
    ungroup %>% 
    mutate(
      mosite = factor(mosite), 
      mosite = factor(mosite, levels = rev(levels(mosite)))
    )
  
  # difference tiled plot
  pdif <- ggplot(toplorev, aes(x = Year, y = mosite, fill = flo_df)) + 
    geom_tile(colour = 'black') + 
    pbase +
    theme(
      panel.background = element_rect(fill = 'lightgrey'), 
      axis.text.y = element_blank()
      ) +
    scale_x_continuous('', expand = c(0, 0)) + 
    scale_y_discrete('', expand = c(0, 0)) +
    scale_fill_gradient2('Difference from average,\nby location', low = 'tomato1', mid = "white", high = 'lightgreen', midpoint = 0) +
    guides(fill = guide_colourbar(barheight = 0.5, barwidth = 5, label.theme = element_text(size = 10, angle = 0, family = 'serif')))
  
  pall + pdif + plot_layout(ncol = 2, widths = c(0.6, 1))
  
}, height = 300)
```

## Fish

```{r}
wellPanel(style = "overflow-y:scroll; max-height: 500px",
renderPlot({
  
  # input
  fshyrs <- fshyrs()
  
  toplo <- fshyrs %>% 
    dplyr::select(-tnd) %>% 
    unnest
  
  pall <- ggplot(toplo, aes(x = SampleDate, y = density, fill = density_df)) + 
    geom_hline(aes(yintercept = avedens), colour = 'grey') +
    geom_line() + 
    geom_point(pch = 21, colour = 'black', size = 3) + 
    geom_smooth(method = 'lm', se = F, linetype = 'dashed', colour = 'black') +
    facet_grid(SiteID ~ .) +
    pbase +
    theme(
      axis.title.x = element_blank(),
      legend.position = 'none',
      axis.text.y = element_text(size = 8)
      ) +
    scale_y_continuous('Density (fish/100 ft)') + 
    scale_fill_gradient2('Difference from average,\nby location', low = 'tomato1', mid = "white", high = 'lightgreen', midpoint = 0) +
    guides(fill = guide_colourbar(barheight = 0.5, barwidth = 5, label.theme = element_text(size = 10, angle = 0)))
  
  toplorev <- toplo %>% 
    ungroup %>% 
    mutate(
      SiteID = factor(SiteID),
      SiteID = factor(SiteID, levels = rev(levels(SiteID)))
    )
  
  # difference tiled plot
  pdif <- ggplot(toplorev) + 
    geom_tile(aes(x = Year, y = SiteID, fill = density_df), colour = 'black') + 
    pbase +
    theme(
      panel.background = element_rect(fill = 'lightgrey'),
      axis.text.y = element_blank()
      ) +
    scale_x_continuous('', expand = c(0, 0)) + 
    scale_y_discrete('', expand = c(0, 0)) +
    scale_fill_gradient2('Difference from average,\nby site', low = 'tomato1', mid = "white", high = 'lightgreen', midpoint = 0) +
    guides(fill = guide_colourbar(barheight = 0.5, barwidth = 5, label.theme = element_text(size = 10, angle = 0, family = 'serif'))) 
  
  pall + pdif + plot_layout(ncol = 2, widths = c(0.6, 1))
  
}, height = 1500)
)
```

# Summary trends {.tabset}

These plots show the estimated trend and significance for flow and fish density estimates at the sites.  Trends are based on a Kendall test for the time series at each site that evaluates the magnitude, direction, and significance of a change between the selected years. The value for tau ranges from -1 to 1 and provides a measure of trend direction. Bars for each site are outlined in black if the trend is significant, with the appropriate text indication above/below each bar (ns = not significant, \* p < 0.05, \* p < 0.005).

## Flow

`r renderText({flotxt()})`
```{r}
renderPlot({
  
  # input
  floyrs <- floyrs()

  toplo <- floyrs %>% 
    dplyr::select(mosite, tnd) %>% 
    unnest %>% 
    arrange(tau) %>% 
    mutate(
      mosite = factor(mosite, levels = mosite), 
      vjsval = ifelse(tau < 0, 1.5, -0.5)
    )
  
  p <- ggplot(toplo, aes(x = mosite, y = tau)) + 
    geom_bar(stat = 'identity', aes(colour = pval, fill = tau), size = 1) + 
    geom_text(aes(label = pval), vjust = toplo$vjsval, size = 6, family = 'serif') + 
    scale_fill_gradient2(low = 'tomato1', mid = "white", high = 'lightgreen', midpoint = 0, limits = c(-1, 1)) +
    scale_colour_manual(values = c(`ns` = NA, `*` = 'black', `**` = 'black'), drop = F, guide = F) + 
    pbase +
    geom_hline(yintercept = 0) + 
    theme(
      axis.title.x = element_blank(),
      legend.position = 'none'
      ) +
    scale_y_continuous(limits = c(-1, 1)) +
    ylab('Trend (tau)') 
    
  return(p)

}, height = 500)
```

## Fish

`r renderText({fshtxt()})`
```{r}
renderPlot({
  
  # input
  fshyrs <- fshyrs()

  toplo <- fshyrs %>% 
    dplyr::select(SiteID, tnd) %>% 
    unnest %>% 
    arrange(tau) %>% 
    mutate(
      SiteID = factor(SiteID, levels = SiteID), 
      vjsval = ifelse(tau < 0, 1.5, -0.5)
    )
  
  p <- ggplot(toplo, aes(x = SiteID, y = tau)) + 
    geom_bar(stat = 'identity', aes(colour = pval, fill = tau), size = 1) + 
    geom_text(aes(label = pval), vjust = toplo$vjsval, size = 6, family = 'serif') + 
    scale_fill_gradient2(low = 'tomato1', mid = "white", high = 'lightgreen', midpoint = 0, limits = c(-1, 1)) +
    scale_colour_manual(values = c(`ns` = NA, `*` = 'black', `**` = 'black'), drop = F, guide = F) + 
    pbase +
    geom_hline(yintercept = 0) + 
    theme(
      axis.title.x = element_blank(),
      legend.position = 'none'
      ) +
    scale_y_continuous(limits = c(-1, 1)) +
    ylab('Trend (tau)')
    
  return(p)

}, height = 500)
```

# By site {.tabset}

These plots show the trends at each individual site and provide a closer view of the plots in the overall trends section. The top plot shows the observed time series with a grey line indicating the overall average. The bottom plot shows the differnece from the average, with a linear trend fit through the values.  Points and bars are colored by difference from the average. 

## Flow

```{r}
column(2, NULL)
column(4, selectInput('sitflo', 'Select flow site:', choices = unique(pajflo$mosite)))
column(6, NULL)
```

```{r}
renderPlot({
  
  # input
  floind <- floind()
  yrssel <- input$yrssel
  
  toplo <- floind 

  p1 <- ggplot(toplo, aes(x = date, y = flo)) + 
    geom_hline(aes(yintercept = aveflo), color = 'darkgrey') + 
    geom_line() +
    geom_point(aes(fill = flo_df), colour = 'black', pch = 21, size = 6) + 
    theme_bw(base_family = 'serif', base_size = 14) +
    theme(
      strip.background = element_blank(),
      panel.grid = element_blank(), 
      legend.position = 'none'
    ) + 
    scale_y_log10('log-Flow (cfs)') + 
    scale_x_date(limits = ymd(paste0(c(yrssel[1], yrssel[2] + 1), c('-01-01')))) +
    scale_fill_gradient2(low = 'tomato1', mid = "white", high = 'lightgreen', midpoint = 0) +
    xlab('Date')
  
  p2 <- ggplot(toplo, aes(x = Year, y = flo_df, fill = flo_df)) + 
    geom_bar(stat = 'identity', colour = 'grey') + 
    scale_fill_gradient2('', low = 'tomato1', mid = 'white', high = 'lightgreen', midpoint = 0) +
    geom_smooth(method = 'lm', se = F, linetype = 'dashed', color = 'black') + 
    theme_bw(base_family = 'serif', base_size = 14) +
    theme(
      strip.background = element_blank(),
      panel.grid = element_blank()
    ) + 
    scale_y_continuous('Flow difference from average') +
    scale_x_continuous(limits = yrssel) +
    geom_hline(yintercept = 0)
  
  p1 + p2 + plot_layout(ncol = 1)

}, height = 600, width = 800)
```

## Fish

```{r}
column(2, NULL)
column(4, selectInput('sitfsh', 'Select fish sample site:', choices = unique(pajfsh$SiteID)))
column(6, NULL)
```

```{r}
renderPlot({
  
  # input
  fshind <- fshind()
  yrssel <- input$yrssel
  
  toplo <- fshind 

  p1 <- ggplot(toplo, aes(x = SampleDate, y = density)) + 
    geom_hline(aes(yintercept = avedens), color = 'darkgrey') + 
    geom_line() +
    geom_point(aes(fill = density_df), pch = 21, colour = 'black', size = 6) + 
    theme_bw(base_family = 'serif', base_size = 14) +
    theme(
      strip.background = element_blank(),
      panel.grid = element_blank(), 
      legend.position = 'none'
    ) + 
    scale_x_date(limits = ymd(paste0(c(yrssel[1], yrssel[2] + 1), c('-01-01')))) +
    scale_fill_gradient2(low = 'tomato1', mid = "white", high = 'lightgreen', midpoint = 0) +
    scale_y_continuous('Density (fish/100ft)')
  
  p2 <- ggplot(toplo, aes(x = Year, y = density_df, fill = density_df)) + 
    geom_bar(stat = 'identity', colour = 'grey') + 
    scale_fill_gradient2('', low = 'tomato1', mid = 'white', high = 'lightgreen', midpoint = 0) +
    geom_smooth(method = 'lm', se = F, linetype = 'dashed', color = 'black') + 
    theme_bw(base_family = 'serif', base_size = 14) +
    theme(
      strip.background = element_blank(),
      panel.grid = element_blank()
    ) + 
    scale_x_continuous(limits = yrssel) +
    scale_y_continuous('Density difference from average') +
    geom_hline(yintercept = 0)
  
  p1 + p2 + plot_layout(ncol = 1)

}, height = 600, width = 800)
```

# Combined fish and flow trends

Fish densities and flow estimates were combined to explore associations between the two. Fish sample sites were matched to the flow records (September only) as follows:

```{r}
totab <- pajbth %>% 
  dplyr::select(Group, SiteID, Site) %>% 
  unique
row.names(totab) <- 1:nrow(totab)
knitr::kable(totab)
```

```{r}
column(12, 
  column(width = 4, 
    pickerInput("pajgrp", label = 'Choose site location group:', choices = unique(fishmtch$Group),
                                                selected = unique(fishmtch$Group),
                                                multiple = TRUE
                                    )
  )
)
```


`r renderText({paste0('Showing trends for ' , input$yrssel[1], ' to ', input$yrssel[2], '.')})`
```{r}
renderPlot({
  
  # input
  pajbthyrs <- pajbthyrs()
  
  req(nrow(pajbthyrs) > 0)
  
  p <- ggplot(pajbthyrs, aes(x = flo, y = density)) + 
    geom_point() + 
    geom_smooth(method = 'lm', se = F, linetype = 'dashed') + 
    scale_x_log10('log-Flow (cfs)') + 
    scale_y_log10('log-Density (fish/100ft)') +
    facet_wrap(~`Size class`, scales = 'free_y') + 
    pbase
  
  return(p)
  
})
```

<br>
```{r, results = 'asis'}
renderUI({
  HTML(stargazer(pajtndmod(), type = 'html', column.labels = c('S1', 'S2'),
            dep.var.labels = c('Density (fish/100ft)')))
  })
```
<br>


