---
title: "Reach habitat analysis"
runtime: shiny
output: 
  html_document:
    toc: true
    toc_float: 
       collapsed: false
---

This set of analyses evaluates differences in habitat measurements at approximate 1/2 mile observations along reach segments.  The habitat data in this section were collected at fixed distances along reaches and are not identified by sites where fish sampling occurred.  As such, the information in this section is useful to identify changes at a high spatial resolution along reaches independent of the fisheries data collected at individual sites.  Habitat changes over time can also be evaluated for an individual reach.

```{r setuphaban, include = FALSE}
# globals
library(knitr)
opts_chunk$set(echo = F, message = F, warning = F)

library(tidyverse)
library(sf)
library(mapview)
library(shiny)
library(scales)
library(leaflet)
library(shinyWidgets)
library(multcompView)
library(RColorBrewer)

source('R/funcs.R')

data(reach)
data(stream)
data(rchdat)

# color vector
cols <- brewer.pal(12, 'Paired')

# summary statistics choice list
sttls <- list(`Between groups, within years` = 'Year', `Between years, within groups` = 'clas')
```

```{r reactives}
# selected reach data
rchsel <- reactive({
  
  # input
  rchid <- input$rchid
  
  # filter reach data, get cumulative and midpoint reach distances by sample year
  out <- rchdat %>% 
    filter(ReachID %in% rchid) %>% 
    arrange(Year, `Hab. #`) %>% 
    group_by(Year) %>% 
    mutate(
      cmlen = cumsum(`Mean length`),
      mdlen = c(cmlen[1] / 2, cmlen[-length(cmlen)] + c(diff(cmlen) / 2))
    ) %>% 
    ungroup
  
  return(out)
  
})

# map with selected sf reach object
spasel <- reactive({
  
  # input
  rchid <- input$rchid
  wshd <- input$wshd
  
  # subset sf by selected watershed
  wshdflt <-  reach %>% 
    filter(Watershed %in% wshd) 
    
  # subset sf by selected reach
  rchflt <- reach %>% 
    filter(ReachID %in% rchid)
  
  # base map
  out <- mapview(reach, label = reach$ReachID, homebutton = F, lwd = 1) +
    mapview(wshdflt, label = wshdflt$ReachID, homebutton = F, lwd = 3) 
  
  # add selected reach if on map
  if(nrow(rchflt) > 0)
    out <- out +
      mapview(rchflt, label = rchflt$ReachID, lwd = 5, color = 'tomato1')
    
  return(out)
  
})

# which habitat variables to select from reactive input clas
hbshsel <- reactive({

  # input
  clas <- input$clas
  rchsel <- rchsel()
  
  req(nrow(rchsel) > 0)

  # get habitat types to show
  tosel <- rchsel %>% 
    rename(
      clas = !!clas
    ) %>% 
    pull(clas) %>% 
    unique %>% 
    sort
  
  return(tosel)
    
})

# color vector for habitat selection
colcat <- reactive({
  
  # input
  hbshsel <- hbshsel()

  # map the colors
  if(length(hbshsel) <= length(cols)){
    colvls <- cols[seq_along(hbshsel)]
  } else {
    colvls <- scales::seq_gradient_pal(cols)(seq(0, 1, length.out = length(hbshsel)))
  }
  
  return(colvls)
  
})
  
# reach data to plot
rchtpl <- reactive({
  
  # input 
  rchsel <- rchsel()
  hbshsel <- hbshsel()
  clas <- input$clas
  szas <- input$szas
  hbsh <- input$hbsh
  ylim <- input$ylim
  
  req(nrow(rchsel) > 0)
  
  # filter by habitat selection, reach distances
  out <- rchsel %>% 
    rename(
      szas = !!szas,
      clas = !!clas
    ) %>% 
    dplyr::select(Year, clas, szas, mdlen) %>% 
    mutate(clas = factor(clas, levels = hbshsel)) %>% 
    filter(clas %in% hbsh) %>% 
    filter(mdlen < ylim[2] & mdlen > ylim[1]) %>% 
    filter(!is.na(szas))
  
  return(out)

})
  
# summary stats
rchstt <- reactive({
  
  # input
  rchtpl <- rchtpl()
  hbshsel <- hbshsel()
  sttgrp <- input$sttgrp
  
  req(nrow(rchtpl) > 0)

  # categorical variable, not for grouping 
  catvr <- unlist(sttls)[!unlist(sttls) %in% sttgrp]
  
  # rename grouping and category columns by inputs
  toout <- rchtpl %>%
    rename(
      grpvr = !!sttgrp,
      catvr = !!catvr
    )

  # stats summary output
  out <- toout %>% 
    group_by(grpvr) %>% 
    nest %>%
    mutate(
      lets = purrr::map(data, function(x){

        # try to get multcomp
        lets <- try({

          anomod <- aov(szas ~ catvr, data = x)
          modhsd <- TukeyHSD(anomod) %>% 
            .$catvr %>% 
            data.frame
          pvals <- modhsd$p.adj
          names(pvals) <- rownames(modhsd)
        
          letout <- multcompLetters(pvals)
          
          letout
          
          })
        
        # run check if possible to do multcomp
        # if multcomp error, make comparable output but with NA        
        if(inherits(lets, 'try-error')){
         
          unicat <- x %>% 
            pull(catvr) %>% 
            unique %>% 
            sort
          lets <- rep(NA, times = length(unicat))
          names(lets) <- unicat
           
        }
        
        if('Letters' %in% names(lets))
          lets <- lets$Letters
        lets <- lets %>% 
          enframe('catvr', 'lets')
        
        return(lets)
        
      })
    ) %>% 
    dplyr::select(-data) %>% 
    unnest() 
    
  # ymax by year and group for label placement on plot
  ymx <- toout %>% 
    group_by(grpvr, catvr) %>% 
    summarise(ymax = max(szas, na.rm = T))
  
  # combine ymx locations with out
  # reassign factor levels to habitat types correctly
  
  if(sttgrp == 'clas')
    out <- out %>% 
      mutate(grpvr = factor(grpvr, levels = hbshsel)) %>% 
      left_join(ymx, by = c('grpvr', 'catvr'))
  if(sttgrp == 'Year')
    out <- out %>% 
      mutate(catvr = factor(catvr, levels = hbshsel)) %>% 
      left_join(ymx, by = c('grpvr', 'catvr'))

  
  # revert names
  names(out)[names(out) == 'grpvr'] <- sttgrp
  names(out)[names(out) == 'catvr'] <- catvr

  return(out)

})

# text for test interpretation
tstint <- reactive({
  
  # inputs
  rchtpl <- rchtpl()
  rchid <- input$rchid
  szas <- input$szas
  clas <- input$clas
  hbsh <- input$hbsh
  sttgrp <- input$sttgrp
  ylim <- input$ylim
  
  # concatenate strings
  yrs <- rchtpl %>% 
    pull(Year) %>% 
    unique %>% 
    sort %>% 
    paste(collapse = ', ')
  hbsh <- hbsh[nchar(hbsh) > 0] %>%
    paste(collapse = ', ')

  # make text output
  
  if(sttgrp == 'Year')
    txt <- paste0('Is ', szas , ' significantly different between habitat selections (', clas , ': ',      hbsh, ') for the same year (', yrs ,') at reach ', rchid , ' (', ylim[1], ' m downstream to ', ylim[2], ' m upstream)?')
                  
  if(sttgrp == 'clas')
    txt <- paste0('Is ', szas , ' significantly different between years (', yrs ,') for the same habitat selection (', clas , ': ',      hbsh, ') at reach ', rchid , ' (', ylim[1], ' m downstream to ', ylim[2], ' m upstream)?')

  return(txt)
  
})

# reach plot data, points
output$rchplopts <- renderPlot({
  
  # input 
  rchtpl <- rchtpl()
  rchstt <- rchstt()
  colcat <- colcat()
  szrn <- input$szrn
  szas <- input$szas
  clas <- input$clas
  ylim <- input$ylim

  # changes over time by dist
  p <- ggplot(rchtpl, aes(x = Year)) + 
    geom_point(aes(y = mdlen, size = szas, fill = clas), colour = 'black', pch = 21, alpha = 0.9)+ 
    ylab('Reach distance (m, downstream to upstream)') + 
    scale_y_continuous(limits = ylim) +
    scale_fill_manual(clas, values = colcat, drop = F) + 
    scale_size(szas, range = c(szrn[1], szrn[2])) + 
    theme_bw(base_family = 'serif', base_size = 16) + 
    theme(
      axis.title.x = element_blank(), 
      legend.position = 'top', 
      legend.box = 'vertical'
      ) + 
    guides(fill = guide_legend(override.aes = list(size = 5)))

  return(p)
  
}, height = 450)

# reach plot data, boxplots
output$rchplobox <- renderPlot({
  
  # input 
  rchtpl <- rchtpl()
  rchstt <- rchstt()
  colcat <- colcat()
  szrn <- input$szrn
  szas <- input$szas
  clas <- input$clas
  
  # boxplot by time
  p <- ggplot(rchtpl, aes(x = Year, fill = clas, y = szas)) + 
    geom_boxplot() +
    ylab(szas) + 
    geom_text(data = rchstt, aes(x = Year, y = ymax, colour = clas, label = lets), position = position_dodge(width = 0.75), size = 6, vjust = -0.5) +
    scale_fill_manual(clas, values = colcat, drop = F) + 
    scale_colour_manual(clas, values = colcat, drop = F, guide = F) +
    theme_bw(base_family = 'serif', base_size = 16) + 
    theme(
      axis.title.x = element_blank(), 
      legend.position = 'top'
      ) 
  
  return(p)
  
}, height = 400)
```

## Reach selection map

Choose the watershed and reach for evaluating habitat changes.  The watershed selection in the first menu determines which reaches are available to select in the second menu. 

```{r}
column(12,
column(2, NULL),
column(4, 
  selectInput("wshd", "Choose watershed:", choices = levels(rchdat$Watershed), selected = 'SLR-main')
),
column(4, 
  renderUI({
    
    # input
    wshd <- input$wshd

    tosel <- rchdat %>% 
      filter(Watershed %in% wshd) %>%
      pull(ReachID) %>% 
      unique %>% 
      sort
    
    selectInput('rchid', "Choose reach:", choices = tosel)
    # pickerInput('rchid', "Choose reach:", choices = tosel, multiple = T, selected = tosel[1])
    
  })
),
column(2, NULL)
)
```

```{r}
renderLeaflet({spasel()@map})
```

## Habitat selection

*Results below are for reach `r renderText({input$rchid})`.*

Habitat data for the selected reach above is shown in the plots below.  Choose the habitat classification for the color aesthetic and grouping variable in the plots with the first menu. Choose the habitat types to show in the second menu based on the habitat classification from the first menu.  The habitat variable that is shown for each habitat type can be chosen with the third menu.  The chosen habitat variable defines how large the points are in the first plot and which variable is shown on the y-axis for the boxplots in the second plot.  

For the selected reach, the slider can further filter the data within the measured distances that were sampled along the reach. The distances start at zero from the first downstream sample site to the maximum sampled distance of a reach across the years. The default position is to plot all the available data along the reach.  These distances can be shortened to view the data along a subset of the total distance that was sampled for the reach.  The selected distances affect plots in both tabs. 

```{r}
column(12,
column(4, 
  selectInput("clas", "Choose color aesthetic:", choices = c('Gen Hab Type', 'Hab. type', 'wood'), selected = 'Gen Hab Type')
),
column(4, 
  renderUI({
    
    # input
    hbshsel <- hbshsel()
    
    req(hbshsel)
    
    pickerInput('hbsh', 'Choose habitat selections to show:', choices = hbshsel, multiple = T, selected = hbshsel,
      options = list(`actions-box` = TRUE)
      )
    
  })     
),
column(4, 
  selectInput("szas", "Choose variable for size or boxplots:", choices = c("mean width", "mean depth", "max depth", "avg. embedd.", "escape cover", "% fines", "% shade", "% deciduous", "cover/ length", "cover/ perimeter"), selected = 'mean width')
  )
)
column(12,
column(8,
  renderUI({

    # input
    rchsel <- rchsel()
    
    req(nrow(rchsel) > 0)

    maxy <- max(rchsel$mdlen, na.rm = TRUE)
        
    sliderInput("ylim", label = "Choose reach distance for both plots (m, downstream to upstream):", min = 0,
          max = maxy, value = c(0, maxy), width = '100%')

  })
)
)
```

## {.tabset}

### Habitat data by sample distance, over time

This plot shows the reach data across years from downstream (bottom of y-axis) to upstream (top of y-axis).  Points are colored by the habitat selections and sized by the chosen habitat variable.  The slider can be used to change the scaling of the point sizes to better distinguish changes along the reach and between years.  

```{r}
column(12, 
column(4, 
  sliderInput("szrn", label = "Choose point size ranges:", min = 0, 
        max = 15, value = c(1, 8))
)
)
plotOutput('rchplopts')
```

### Habitat data distribution, over time

This plot shows the distribution of the selected habitat variable as boxplots for each year, grouped and colored by the habitat selections.  Summary statistics are also shown based on the chosen comparison in the menu below.  The summary statistics can be used to evaluate differences __between habitat selections within the same year__ or __between years within the same habitat selection__.  Boxplots that share the same letter are not statistically different (at alpha = 0.05, p-values corrected for multiple comparisons). The comparison of letters are only relevant for the selected summary statistics from the drop down menu. That is, letters can be only be compared within the same year if "between groups, within years" is selected, or letters can only be compared between years for the same group (habitat selection) if "between years, within groups" is selected.  Comparisons for boxplots without letters are not shown if the test could not be completed due to sample size.

```{r}
column(12, 
column(8, 
  selectInput('sttgrp', label = "Choose summary statistics:", choices = sttls)
)       
)
```

*This test asks the question:*

`r renderText({tstint()})`
```{r}
plotOutput('rchplobox')
```




