---
title: "Reach habitat analysis"
runtime: shiny
output: 
  html_document:
    toc: true
    toc_float: 
       collapsed: false
---

This set of analyses evaluates differences in habitat measurements at approximate 1/2 mile observations along reach segments.  The habitat data in this section were collected at fixed distances along reaches and are not identified by sites where fish sampling occurred.  As such, the information in this section is useful to identify changes at a high spatial resolution along reaches independent of the fisheries data collected at individual sites.  Habitat changes over time can also be evaluated for an individual reach.

```{r setuphaban, include = FALSE}
# globals
library(knitr)
opts_chunk$set(echo = F, message = F, warning = F)

library(tidyverse)
library(sf)
library(mapview)
library(shiny)
library(scales)
library(leaflet)
library(shinyWidgets)
library(multcompView)

source('R/funcs.R')

data(reach)
data(stream)
data(rchdat)

# summary statistics choice list
sttls <- list(`Within years, between groups` = 'Year', `Within groups, between years` = 'clas')
```

```{r reactives}
# selected reach data
rchsel <- reactive({
  
  # input
  rchid <- input$rchid
  
  # filter reach data, get cumulative and midpoint reach distances by sample year
  out <- rchdat %>% 
    filter(ReachID %in% rchid) %>% 
    arrange(Year, `Hab. #`) %>% 
    group_by(Year) %>% 
    mutate(
      cmlen = cumsum(`Mean length`),
      mdlen = c(cmlen[1] / 2, cmlen[-length(cmlen)] + c(diff(cmlen) / 2))
    ) %>% 
    ungroup
  
  return(out)
  
})

# map with selected sf reach object
spasel <- reactive({
  
  # input
  rchid <- input$rchid
  wshd <- input$wshd
  
  # subset sf by selected watershed
  wshdflt <-  reach %>% 
    filter(Watershed %in% wshd) 
    
  # subset sf by selected reach
  rchflt <- reach %>% 
    filter(ReachID %in% rchid)
  
  # base map
  out <- mapview(reach, label = reach$ReachID, homebutton = F, lwd = 1) +
    mapview(wshdflt, label = wshdflt$ReachID, homebutton = F, lwd = 3) 
  
  # add selected reach if on map
  if(nrow(rchflt) > 0)
    out <- out +
      mapview(rchflt, label = rchflt$ReachID, lwd = 5, color = 'tomato1')
    
  return(out)
  
})

# reach data to plot
rchtpl <- reactive({
  
  # input 
  rchsel <- rchsel()
  clas <- input$clas
  szas <- input$szas
  hbsh <- input$hbsh
  ylim <- input$ylim
  
  req(nrow(rchsel) > 0)

  # get midpoint of habitat sample on reach
  out <- rchsel %>% 
    rename(
      szas = !!szas,
      clas = !!clas
    ) %>% 
    dplyr::select(Year, clas, szas, mdlen) %>% 
    filter(clas %in% hbsh) %>% 
    filter(mdlen < ylim[2] & mdlen > ylim[1]) %>% 
    filter(!is.na(szas))
  
  return(out)

})
  
# summary stats
rchstt <- reactive({
  
  # input
  rchtpl <- rchtpl()
  sttgrp <- input$sttgrp
  
  req(nrow(rchtpl) > 0)
  
  # categorical variable, not for grouping 
  catvr <- unlist(sttls)[!unlist(sttls) %in% sttgrp]
  
  # rename grouping and category columns by inputs
  toout <- rchtpl %>%
    rename(
      grpvr = !!sttgrp,
      catvr = !!catvr
    )
  
  # stats summary output  
  out <- toout %>% 
    group_by(grpvr) %>% 
    nest %>%
    mutate(
      anomod = purrr::map(data, function(x){
        
        aov(szas ~ catvr, data = x)
        
      }), 
      lets = purrr::map(anomod, function(x){
        
        modhsd <- TukeyHSD(x) %>% 
          .$catvr %>% 
          data.frame
        pvals <- modhsd$p.adj
        names(pvals) <- rownames(modhsd)
        lets <- multcompLetters(pvals)
        if('Letters' %in% names(lets))
          lets <- lets$Letters
        lets <- lets %>% 
          enframe('catvr', 'lets')
        
      })
    ) %>% 
    dplyr::select(-data, -anomod) %>% 
    unnest() 
    
  # ymax by year and group for label placement on plot
  ymx <- toout %>% 
    group_by(grpvr, catvr) %>% 
    summarise(ymax = max(szas, na.rm = T))
  
  # combine ymx locations with out
  out <- out %>% 
    left_join(ymx, by = c('grpvr', 'catvr'))

  # revert names
  names(out)[names(out) == 'grpvr'] <- sttgrp
  names(out)[names(out) == 'catvr'] <- catvr

  return(out)

})

# reach plot data, points
output$rchplopts <- renderPlot({
  
  # input 
  rchtpl <- rchtpl()
  rchstt <- rchstt()
  szrn <- input$szrn
  szas <- input$szas
  clas <- input$clas

  # changes over time by dist
  p <- ggplot(rchtpl, aes(x = Year)) + 
    geom_point(aes(y = mdlen, size = szas, fill = clas), colour = 'black', pch = 21, alpha = 0.9)+ 
    ylab('Reach distance (m, downstream to upstream)') + 
    scale_fill_brewer(clas, palette = 'Paired') + 
    scale_size(szas, range = c(szrn[1], szrn[2])) + 
    theme_bw(base_family = 'serif', base_size = 16) + 
    theme(
      axis.title.x = element_blank(), 
      legend.position = 'top', 
      legend.box = 'vertical'
      ) + 
    guides(fill = guide_legend(override.aes = list(size = 5)))

  return(p)
  
}, height = 450)

# reach plot data, boxplots
output$rchplobox <- renderPlot({
  
  # input 
  rchtpl <- rchtpl()
  rchstt <- rchstt()
  szrn <- input$szrn
  szas <- input$szas
  clas <- input$clas
  
  # boxplot by time
  p <- ggplot(rchtpl, aes(x = Year, fill = clas, y = szas)) + 
    geom_boxplot() +
    ylab(szas) + 
    geom_text(data = rchstt, aes(x = Year, y = ymax, colour = clas, label = lets), position = position_dodge(width = 0.75), size = 6, vjust = -0.5) +
    scale_fill_brewer(clas, palette = 'Paired') + 
    scale_colour_brewer(clas, palette = 'Paired', guide = 'none') +
    theme_bw(base_family = 'serif', base_size = 16) + 
    theme(
      axis.title.x = element_blank(), 
      legend.position = 'top'
      ) 
  
  return(p)
  
}, height = 400)
```

## Reach selection map

Choose the watershed and reach for evaluating habitat changes.  The watershed selection in the first menu determines which reaches are available to select in the second menu. 

```{r}
column(12,
column(2, NULL),
column(4, 
  selectInput("wshd", "Choose watershed:", choices = levels(rchdat$Watershed), selected = 'SLR-main')
),
column(4, 
  renderUI({
    
    # input
    wshd <- input$wshd

    tosel <- rchdat %>% 
      filter(Watershed %in% wshd) %>%
      pull(ReachID) %>% 
      unique %>% 
      sort
    
    selectInput('rchid', "Choose reach:", choices = tosel)
    # pickerInput('rchid', "Choose reach:", choices = tosel, multiple = T, selected = tosel[1])
    
  })
),
column(2, NULL)
)
```

```{r}
renderLeaflet({spasel()@map})
```

## Habitat selection

Habitat data for the selected reach above is shown in the plots below.  Choose the habitat classification for the color aesthetic and grouping variable in the plots with the first menu. Choose the habitat types to show in the second menu based on the habitat classification from the first menu.  The habitat variable that is shown for each habitat type can be chosen with the third menu.  The chosen habitat variable defines how large the points are in the first plot and which variable is shown on the y-axis for the boxplots in the second plot.  

For the selected reach, the slider can further filter the data within the measured distances that were sampled along the reach. The distances start at zero from the first downstream sample site to the maximum sampled distance of a reach across the years. The default position is to plot all the available data along the reach.  These distances can be shortened to view the data along a subset of the total distance that was sampled for the reach.  The selected distances affect plots in both tabs. 

```{r}
column(12,
column(4, 
  selectInput("clas", "Choose color aesthetic:", choices = c('Gen Hab Type', 'Hab. type', 'wood'), selected = 'Gen Hab Type')
),
column(4, 
  renderUI({
    
    # input
    clas <- input$clas
    rchsel <- rchsel()
  
    req(nrow(rchsel) > 0)

    # get habitat types to show
    tosel <- rchsel %>% 
      rename(
        clas = !!clas
      ) %>% 
      pull(clas) %>% 
      unique
    
    pickerInput('hbsh', 'Choose habitat selections to show:', choices = tosel, multiple = T, selected = tosel,
      options = list(`actions-box` = TRUE)
      )
    
  })     
),
column(4, 
  selectInput("szas", "Choose variable for size or boxplots:", choices = c("mean width", "mean depth", "max depth", "avg. embedd.", "escape cover", "% fines", "% shade", "% deciduous", "cover/ length", "cover/ perimeter"), selected = 'mean width')
  )
)
column(12,
column(8,
  renderUI({

    # input
    rchsel <- rchsel()
    
    req(nrow(rchsel) > 0)

    maxy <- max(rchsel$mdlen, na.rm = TRUE)
        
    sliderInput("ylim", label = "Choose reach distance for both plots (m, downstream to upstream):", min = 0,
          max = maxy, value = c(0, maxy), width = '100%')

  })
)
)
```

## {.tabset}

### Habitat data by sample distance, over time

This plot shows the reach data across years from downstream (bottom of y-axis) to upstream (top of y-axis).  Points are colored by the habitat selections and sized by the chosen habitat variable.  The slider can be used to change the scaling of the point sizes to better distinguish changes along the reach and between years.  

```{r}
column(12, 
column(4, 
  sliderInput("szrn", label = "Choose point size ranges:", min = 0, 
        max = 15, value = c(1, 8))
)
)
plotOutput('rchplopts')
```

### Habitat data distribution, over time

This plot shows the distribution of the selected habitat variable as boxplots for each year, grouped and colored by the habitat selections.  Summary statistics are also shown based on the chosen comparison in the menu below.  The summary statistics can be used to evaluate differences __within the same year but between groups__ (the habitat selection) or __within the same group but between years__.  Boxplots that share the same letter are not statistically different (at alpha = 0.05). The comparison of letters are only relevant for the selected summary statistics from the drop down menu. That is, letters can be only be compared within the same year if "within years, between groups" is selected, or letters can only be compared between years for the same group (habitat selection) if "within groups, between years" is selected.

```{r}
column(12, 
column(8, 
  selectInput('sttgrp', label = "Choose summary statistics:", choices = sttls)
)       
)

plotOutput('rchplobox')
```




