---
title: "Salmonid status and trends"
runtime: shiny
output: 
  html_document:
    toc: true
    toc_float: 
       collapsed: false
    number_sections: true
---

```{r setup, include = FALSE}
# globals
library(knitr)
opts_chunk$set(echo = F, message = F, warning = F)

library(tidyverse)
library(sf)
library(mapview)
library(plotly)
library(shiny)
library(scales)
library(leaflet)
library(stargazer)
library(EnvStats)
library(kableExtra)

prj <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"

source('R/funcs.R')

data(fishdat)
data(stream)
data(trndst_prep)
```

The objective of these analyses is to evaluate the status and trends of Steelhead salmon in the four watershed (note that San Lorenzo is divided by mainstem and tributaries):

* SLR-main: San Lorenzo, mainstem
* SLR-trib: San Lorenzo, tributaries
* SOQ: Soquel
* APT: Aptos
* PAJ: Pajaro

Maps and summary plots show the changes in density (fish/100ft) over time of the different size classes (`Dens_S1` or `Dens_S2`). Salmonid density (fish/100ft) at each site is measured as S1 and S2, where S1 is considered young-of-the-year (< 75mm standard length) and S2 is larger individuals that are either fast-growing young-of-the-year or older fish. S3 fish (> 150 mm standard length) are also grouped with S2.  S2 fish are likely to migrate the following spring but may also be resident trout. Trend statistics of the different size classes for aggregated watershed data and by site are also provided.  The trend analyses only evaluate an increase or decrease (monotonic change) between selected years.

Plese visit this website for a storymap overview of steelhead in the study region: <http://scceh.com/steelhead/overview/>

# Overview maps

This map shows the stream network and sample sites by watershed.  Click on a stream reach or sample site for additional information.  All sites with data collected during any of the sample years are shown. 

```{r mapviewshed, out.width="100%"}
mapview(stream, label = stream$STREAM_NM, legend = F, homebutton = F) +
  mapview(fishdat, zcol = 'Watershed', legend = T, label = fishdat$SiteID, layer.name = 'Watershed', homebutton = F)
```

<br></br>

This map shows the density of S1 or S2 individuals for a chosen year where data are available.  Select a year and size class to view on the map.  The size of the point shows the relative density at a sampled site. 

```{r mapviewsel}
column(2, NULL)
column(4, 
  selectInput("cls", "Choose density (S1: <75 mm, S2: >75 mm):", c('Dens_S1', 'Dens_S2'))
)
column(4, 
  selectInput("yr", "Choose a year to plot:", fishdat$Year %>% unique %>% na.omit)
)
column(2, NULL)

mapout <- reactive({
  
  toplo <- fishdat %>% 
    filter(Year %in% input$yr)

  out <- mapview(stream, label = stream$STREAM_NM, legend = F, homebutton = F) +
    mapview(toplo, zcol = 'Watershed', cex = input$cls, label = as.character(toplo[[input$cls]]), legend = T, layer.name = 'Watershed', homebutton = F)
  
  return(out)
  
})

renderLeaflet({mapout()@map})
```



<br></br>

# Overview plots

The boxplots (description [here](https://upload.wikimedia.org/wikipedia/commons/a/ac/Box_plot_description.jpg)) show the distribution of the density estimates across all sites for a given year, by watershed. 
```{r}
column(2, NULL)
column(4, 
  selectInput('shd', 'Select watershed:', levels(fishdat$Watershed))
)

column(4, 
       selectInput('scl', 'Select scale:', c('log10', 'arithmetic'))
       )
column(2, NULL)
```

```{r densbox}
renderPlot({
  
  # inputs
  shd <- input$shd
  scl <- input$scl

  toplo <- fishdat
  st_geometry(toplo) <- NULL
  toplo <- toplo %>% 
    dplyr::select(Year, SiteID, Watershed, Dens_S1, Dens_S2) %>% 
    gather('Size class', 'density', Dens_S1, Dens_S2) %>% 
    mutate(Year = factor(Year)) %>% 
    # filter(!is.na(Year)) %>% 
    filter(Watershed %in% shd)
 
  p <- ggplot(toplo, aes(x = Year, y = density, fill = `Size class`)) + 
    geom_boxplot() + 
    facet_wrap(~Watershed, ncol = 2) + 
    theme_bw(base_family = 'serif', base_size = 18) + 
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1), 
      strip.background = element_blank(), 
      axis.title.x = element_blank(), 
      legend.position = 'top'
    ) + 
    guides(fill = guide_legend(title = 'Size class'))
 
  if(scl == 'log10')
    p <- p + scale_y_continuous("Density (fish/100ft)", trans = 'log10')
  else 
    p <- p + scale_y_continuous("Density (fish/100ft)")
  
  return(p)
        
}, height = 400, width = 900)

```

<br></br>

These two plot show the overall annual average across years (+/- 95% confidence interval) and changes from the long-term average by year for each watershed and size class.  The colors show the magnitude of the average in the top plot and the differences of each year from the long-term average (trend anomaly) in the bottom plot.  The bottom plot can be used to evaluate an approximate change over time relative to the long-term average in the top plot.

```{r denstile, fig.height = 4, fig.width = 7, out.width = "100%"}
# remove 1981 all watersheds
# remove 1994 from PAJ and SAQ
toeval <- fishdat %>% 
  filter(Year > 1981) %>% 
  filter(!(Year == 1994 & Watershed %in% c('PAJ', 'SOQ')))

# get average differences across stations
st_geometry(toeval) <- NULL
toplo <- toeval %>% 
  gather('Size class', 'density', Dens_S1, Dens_S2) %>% 
  mutate(density = density) %>% 
  group_by(Watershed, `Size class`, Year) %>% 
  summarise(
    density = mean(density, na.rm = T)
    ) %>% 
  group_by(Watershed, `Size class`) %>% 
  mutate(
    density_av = mean(density, na.rm = T),
    density_df = density - density_av
  ) %>%
  na.omit

# average diff and ci per wshed and size class
toploave <- toplo %>% 
  select(Watershed, `Size class`, density, density_av) %>% 
  mutate(
    hi = t.test(density)$conf.int[2],
    lo = t.test(density)$conf.int[1]
  ) %>% 
  select(-density) %>% 
  unique
  
pbase <- theme_bw(base_family = 'serif') +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8), 
    axis.text.y = element_text(size = 8),
    legend.position = 'top',
    legend.direction = 'horizontal',
    # plot.margin = unit(c(4,4,0,0), "lines"),
    strip.background = element_blank(), 
    strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5), 
    panel.background = element_rect(fill = 'lightgrey')
  ) 

# ave barplot
pave <- ggplot(toploave, aes(x = Watershed, y = density_av, fill = density_av)) + 
  geom_bar(stat = 'identity', colour = 'lightgrey') + 
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.5) + 
  coord_flip() +
  scale_y_log10('Average density across years (fish/100ft, log-scale)') + #, expand = c(0, 0)) +
  facet_wrap(~`Size class`, ncol = 1) +
  scale_fill_gradient2('Long-term log-average', low = 'tomato1', mid = "white", high = 'lightgreen', midpoint = mean(toploave$density_av)) +
  guides(fill = guide_colourbar(barheight = 0.5, barwidth = 5, label.theme = element_text(size = 6, angle = 0))) +
  pbase +
  theme(
    panel.background = element_rect(fill = 'white'), 
    axis.title.y = element_blank()
    )

# difference tiled plot
pdif <- ggplot(toplo) + 
  geom_tile(aes(x = Year, y = Watershed, fill = density_df), colour = 'black') + 
  pbase +
  scale_x_continuous('', expand = c(0, 0)) + 
  scale_y_discrete('', expand = c(0, 0)) +
  scale_fill_gradient2('Difference from average,\nby watershed, density', low = 'tomato1', mid = "white", high = 'lightgreen', midpoint = 0) +
  guides(fill = guide_colourbar(barheight = 0.5, barwidth = 5, label.theme = element_text(size = 6, angle = 0))) +
  facet_wrap(~ `Size class`, ncol = 1)

pave
pdif
```

<br></br>

# Trend statistics, by watershed {.tabset}

These plots evaluate trends as measured by the significance of the regression through the trend anomalies. The trend anomalies are shown as positive (green) or negative (red) values around the long-term average (black-line) for a selected watershed and year range.  The dashed line is the regression fit through the anomalies.  All sites in a watershed are grouped. The slider can be used to adjust the range of years that are evaluated, e.g., a trend may be observed for one range but not another.

```{r trendevalrct}

##
# reactive inputs for plots, tables 
column(12, 
column(4, NULL),
# year slider
column(4, 
  sliderInput("years", label = 'Select year ranges to test:',  
        min = 1994, max = 2018, 
        value = c(1994, 2018),
        sep = '', ticks = FALSE
      )
),
column(4, NULL)
)

# plot data from year input, s1
toplos1 <- reactive({
  
  toplos1 <- toplo %>% 
    filter(`Size class` %in% 'Dens_S1') %>% 
    filter(Year >= input$years[1] & Year <= input$years[2])
  
  return(toplos1)
  
})

# plot data from year input, s2
toplos2 <- reactive({
  
  toplos2 <- toplo %>% 
    filter(`Size class` %in% 'Dens_S2') %>% 
    filter(Year >= input$years[1] & Year <= input$years[2])
  
  return(toplos2)
  
})

# table data from plot input, s1
totabs1 <- reactive({
  
  totabs1 <- toplos1() %>% 
    group_by(Watershed) %>% 
    nest %>% 
    mutate(
      res = purrr::map(data, function(x){
        
        lmout <- lm(density_df ~ Year, x) 
        return(lmout)
        
      })
  
    ) %>% 
    select(-data)
  
  return(totabs1)
  
})

# table data kendall test from plot 1, s1
totabs1kn <- reactive({
  
  totabs1kn <- toplos1() %>% 
    group_by(Watershed) %>% 
    nest %>% 
    mutate(
      res = purrr::map(data, function(x){
        
        knout <- kendallTrendTest(density_df ~ Year, x)
        outest <- round(knout$estimate, 2)
        outpval <- p_ast(knout$p.value)
        out <- c(outest, pval = outpval) %>% 
          data.frame %>% 
          t %>% 
          data.frame %>% 
          select(-intercept)
        return(out)
        
      })
  
    ) %>% 
    select(-data) %>% 
    unnest
  
  return(totabs1kn)
  
})

# table data from plot input, s2
totabs2 <- reactive({
  
  totabs2 <- toplos2() %>% 
    group_by(Watershed) %>% 
    nest %>% 
    mutate(
      res = purrr::map(data, function(x){
        
        lmout <- lm(density_df ~ Year, x) 
        return(lmout)
        
      })
  
    ) %>% 
    select(-data)
  
  return(totabs2)
  
})

# table data kendall test from plot 1, s2
totabs2kn <- reactive({
  
  totabs2kn <- toplos2() %>% 
    group_by(Watershed) %>% 
    nest %>% 
    mutate(
      res = purrr::map(data, function(x){
        
        knout <- kendallTrendTest(density_df ~ Year, x)
        outest <- round(knout$estimate, 2)
        outpval <- p_ast(knout$p.value)
        out <- c(outest, pval = outpval) %>% 
          data.frame %>% 
          t %>% 
          data.frame %>% 
          select(-intercept)
        return(out)
        
      })
  
    ) %>% 
    select(-data) %>% 
    unnest
  
  return(totabs2kn)
  
})

```

## S1 year class

```{r s1trnd, fig.height = 4, fig.width = 8, out.width = "100%"}
renderPlot({
  
  p <- ggplot(toplos1(), aes(x = Year, y = density_df, fill = density_df)) + 
    geom_bar(stat = 'identity', colour = 'grey') + 
    scale_fill_gradient2('', low = 'tomato1', mid = 'white', high = 'lightgreen', midpoint = 0) +
    facet_wrap(~ Watershed) + 
    geom_smooth(method = 'lm', se = F, linetype = 'dashed', color = 'black') + 
    theme_bw(base_family = 'serif', base_size = 14) +
    theme(
      strip.background = element_blank(),
      panel.grid = element_blank()
    ) + 
    scale_y_continuous('Density (fish/100ft)') +
    geom_hline(yintercept = 0) + 
    ggtitle('S1 size class')
  
  return(p)
  
})
```

<br></br>

This table shows the regression results from the above plots. 

```{r, results = 'asis'}
renderUI({
  HTML(stargazer(totabs1()$res, type = 'html', column.labels = as.character(totabs1()$Watershed),
            dep.var.labels = 'Density (fish/100ft)'))
  })
```

<br></br>

Kendall tests also provide an indication of a trend by evaluating magnitude, direction, and significance of a change in density between the selected years.  The test is a non-parametric equivalent to the regression analysis above.  The value for tau ranges from -1 to 1 and provides a measure of trend direction.  The slope is the estimated change per year in density and the p-value shows the significance of the test. 

```{r, results = 'asis'}
renderUI({
  HTML(knitr::kable(totabs1kn(), format = 'html') %>% 
    kable_styling(full_width = T, font_size = 14))
  })
```

## S2 year class

```{r s2trnd, fig.height = 4, fig.width = 8, out.width = "100%"}
renderPlot({
  
  p <- ggplot(toplos2(), aes(x = Year, y = density_df, fill = density_df)) + 
    geom_bar(stat = 'identity', colour = 'grey') + 
    scale_fill_gradient2('', low = 'tomato1', mid = 'white', high = 'lightgreen', midpoint = 0) +
    facet_wrap(~ Watershed) + 
    geom_smooth(method = 'lm', se = F, linetype = 'dashed', color = 'black') + 
    theme_bw(base_family = 'serif', base_size = 14) +
    theme(
      strip.background = element_blank(),
      panel.grid = element_blank()
    ) + 
    scale_y_continuous('Density (fish/100ft)') +
    geom_hline(yintercept = 0) + 
    ggtitle('S2 size class')
  
  return(p)
  
})
```

<br></br>

This table shows the regression results from the above plots. 

```{r, results = 'asis'}
renderUI({
  HTML(stargazer(totabs2()$res, type = 'html', column.labels = as.character(totabs2()$Watershed),
            dep.var.labels = 'Density (fish/100ft)'))
  })
```

<br></br>

Kendall tests also provide an indication of a trend by by evaluating magnitude, direction, and significance of a change in density between the selected years.  The test is a non-parametric equivalent to the regression analysis above.  The value for tau ranges from -1 to 1 and provides a measure of trend direcdtion.  The slope is the estimated change per year in density and the p-value shows the significance of the test. 

```{r, results = 'asis'}
renderUI({
  HTML(knitr::kable(totabs2kn(), format = 'html') %>% 
    kable_styling(full_width = T, font_size = 14))
  })
```

# Trend statistics, by site

This section evaluates status and trends of salmonid density at individual sites, rather than aggregated by watershed as above.  Select the year range and size density class to evaluate the trends.  The map shows the value for tau (direction of trend) for a Kendall test of density changes within the selected years, green for increasing and red for decreasing.  Size of the point is the magnitude of the estimated change. Note that some sites may not have data spanning the full range of selected years.  Click an individual site on the map to see the available years within the selected range. 

```{r sitetrndmapprp}
##
# reactive inputs for plots, tables 
column(2, NULL)
column(4, 
  selectInput("clsst", "Choose density (S1: <75 mm, S2: >75 mm):", c('Dens_S1', 'Dens_S2'))
)

# year slider
column(4, 
  sliderInput("yearsst", label = 'Select year ranges to test:',  
        min = 1994, max = 2018, 
        value = c(1994, 2018),
        sep = '', ticks = FALSE
      )
)
column(2, NULL)

# filter site data by density class, years
# get kendall results where appropriate
mapdatst <- reactive({
  
  toplo <- trndst_prep %>% 
    filter(`Size class` %in% input$clsst) %>% 
    mutate(trnd = purrr::map(data, function(x){
      
      xsub <- x %>% 
        filter(Year >= input$yearsst[1] & Year <= input$yearsst[2]) %>%
        mutate(
          avedens = mean(density, na.rm = T), 
          density = density - avedens
          ) %>% 
        dplyr::select(-avedens)
      
      # get unique years tested
      yrs <- unique(xsub$Year) %>% 
        paste(collapse = ', ')

      # kendall sum
      knout <- suppressWarnings(try({kendallTrendTest(density ~ Year, xsub)}, silent = T))
      if(inherits(knout, 'try-error')) return(NA)
      kpval <- knout$p.value
      kest <- knout$estimate
      out <- c(kpval, kest) %>% 
        as.list %>% 
        data.frame %>% 
        mutate(yrs = yrs)
         
      return(out)

      })
    ) %>% 
    dplyr::select(-data) %>% 
    filter(map_lgl(trnd, ~ !anyNA(.x))) %>% # this!
    unnest %>% 
    dplyr::select(-intercept) %>% 
    mutate(
      z = p_ast(z), 
      trend = sign(tau),
      trend = factor(trend, levels = c(-1, 1), labels =c('dec', 'inc'))
      ) %>% 
    mutate_if(is.numeric, round, 2) %>% 
    rename(pval = z) %>% 
    st_as_sf(coords = c("X", "Y"), crs =prj)
  
  return(toplo)
  
  })

# create map from selected, trended data
mapoutst <- reactive({

  mapdatst <- mapdatst()
  
  # point colors
  cols <- mapdatst %>% 
    mutate(
      cols = factor(trend, levels = c('dec', 'inc'), labels = c('tomato1', 'lightgreen')),
      cols = as.character(cols)
    ) %>% 
    pull(cols)

  # size values
  cexv <- mapdatst%>% 
    pull(tau) %>% 
    abs %>% 
    scales::rescale(to = c(2, 15))

  # hover pt labels
  labs <- paste(mapdatst$SiteID, ': ', mapdatst$trend, ', tau = ', mapdatst$tau, ', p = ', mapdatst$pval)
  
  out <- mapview(stream, label = stream$STREAM_NM, homebutton = F, legend = F) +
    mapview(mapdatst, zcol = 'trend', cex = cexv, label = labs, col.regions = cols, legend = F, homebutton = F)
  
  return(out)
  
  
  })
```
```{r sitetrndmap}
# render the site trend map  
renderLeaflet({mapoutst()@map})
```

A regression through the trend anomalies for the size class densities at a selected site is shown below.  These plots can confirm the results shown in the map. 

```{r sitetrndpltprp}

column(4, NULL)
# dynamic ui that changes with year selections on map, differents sites depending on year combos
column(4, 
  renderUI({
  
    sts <- mapdatst()$SiteID %>% unique %>% sort
    selectInput("st", "Select station to view map trends:", sts)
  
  })
)
column(4, NULL)


# filter site data by years, individual site
trndatst <- reactive({
  
  yrsel <- input$yearsst
  stsel <- input$st
  
  req(!is.null(stsel))
  
  toplo <- trndst_prep %>% 
    filter(SiteID %in% stsel) %>% 
    unnest %>% 
    filter(Year >= yrsel[1] & Year <= yrsel[2]) %>% 
    group_by(`Size class`) %>% 
    mutate(
      avedens = mean(density, na.rm = T), 
      density_df = density - avedens
      ) %>% 
    dplyr::select(-avedens)
  
  return(toplo)

  })
```

```{r sitetrndplt, fig.height = 4, fig.width = 8, out.width = "100%"}
# plot of year trends at a site
renderPlot({
  
  toplo <- trndatst()
  stsel <- input$st

  req(!is.null(stsel))
  
  p <- ggplot(toplo, aes(x = Year, y = density_df, fill = density_df)) + 
    geom_bar(stat = 'identity', colour = 'grey') + 
    scale_fill_gradient2('', low = 'tomato1', mid = 'white', high = 'lightgreen', midpoint = 0) +
    facet_wrap(~ `Size class`, ncol = 2) + 
    geom_smooth(method = 'lm', se = F, linetype = 'dashed', color = 'black') + 
    theme_bw(base_family = 'serif', base_size = 14) +
    theme(
      strip.background = element_blank(),
      panel.grid = element_blank()
    ) + 
    scale_y_continuous('Density (fish/100ft)') +
    geom_hline(yintercept = 0) + 
    ggtitle(paste0('Trends at ', stsel))
  
  return(p)
  
})

```

```{r test}
plot(rnorm(100))
```
