---
title: "Factors associated with trends"
runtime: shiny
output: 
  html_document:
    toc: true
    toc_float: 
       collapsed: false
---

```{r setup, include = FALSE}
# globals
library(knitr)
opts_chunk$set(echo = F, message = F, warning = F)

library(tidyverse)
library(sf)
library(shiny)
library(scales)
library(leaflet)
library(stargazer)
library(EnvStats)
library(kableExtra)
library(gridExtra)
library(MuMIn)

prj <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"

source('R/funcs.R')

data(fishdat)
data(habitat)
data(allfctprs)
data(alltops)

# combined salmonid, habitat data, only where years intersect
saldat <- fishdat
st_geometry(saldat) <- NULL
saldat <- saldat %>% 
  dplyr::select(Year, SiteID, Watershed, Dens_S1, Dens_S2) %>% 
  group_by(Year, SiteID, Watershed) %>% 
  summarise(
    Dens_S1 = mean(Dens_S1, na.rm = T), 
    Dens_S2 = mean(Dens_S2, na.rm = T)
  ) %>% 
  ungroup %>% 
  mutate(SiteID = as.character(SiteID))

habdat <- habitat
st_geometry(habdat) <- NULL
habdat <- habdat %>% 
  dplyr::select(-Watershed) %>% 
  group_by(Year, SiteID, HabType) %>% 
  filter(!habvar %in% 'StnSthd') %>% 
  nest %>% 
  filter(HabType %in% c('pool', 'riffle', 'run'))

# combind
dat <- inner_join(saldat, habdat, by = c('Year', 'SiteID'))

# habitat variables to select, master obj
habvrs <- list(
  'Canopy cover (%)' = 'StnCan',
  'Deciduous canopy cover (%)' = 'StnDecid', 
  'Average depth (ft)' = 'StnDpthAvg',
  'Maximum depth (ft)' = 'StnDpthMax',
  'Embeddedness (%)' = 'StnEmbed',
  'Escape cover (ratio)' = 'StnEsCov',
  'Fines (%)' = 'StnFines',
  'Station length (ft)' = 'StnLgth',
  'Station width (ft)' = 'StnWdth'
  )
```

The objective of these analyses is to identify habitat factors that are associated with salmonid trends. Trends are evaluated separately for the following:

* Salmonid density size classes (S1 and S2)
* Watershed (APT, PAJ, SLR, SOQ)
* Habitat type (run, riffle, pool)

# Bivariate comparisons, averaged by watershed {.tabset}

These analyses test bivariate comparisons of a selected habitat variable (by habitat type) and year against changes in salmonid density.  Analyses are evaluated by each watershed where all density and habitat estimates are averaged across stations within each year prior to analysis.

```{r inps}
column(4, 
      selectInput('shd', 'Select waterhsed:', sort(unique(dat$Watershed))) 
       )

column(4, 
       selectInput('typ', 'Select habitat type:', sort(unique(dat$HabType)))
       )

column(4, 
       selectInput('hbv', 'Select habitat variable:', habvrs)
       )
```

```{r}
# dens_s1 data averaged for plot, models
dens1_avg <- reactive({
  
  # inputs
  shd <- input$shd
  typ <- input$typ
  hbv <- input$hbv
  
  dens1_avg <- dat %>% 
    unnest %>% 
    filter(HabType %in% typ & habvar %in% hbv & Watershed %in% shd) %>% 
    select(-Dens_S2, -Watershed) %>% 
    group_by(Year, habvar) %>% 
    summarise(
      Dens_S1 = mean(Dens_S1, na.rm = T),
      habval = mean(habval, na.rm = T)
    ) %>% 
    ungroup %>% 
    spread(habvar, habval) 
  
  return(dens1_avg)
  
})

# dens_s2 data averaged for plot, models
dens2_avg <- reactive({
  
  # inputs
  shd <- input$shd
  typ <- input$typ
  hbv <- input$hbv
  
  dens2_avg <- dat %>% 
    unnest %>% 
    filter(HabType %in% typ & habvar %in% hbv & Watershed %in% shd) %>% 
    select(-Dens_S1, -Watershed) %>% 
    group_by(Year, habvar) %>% 
    summarise(
      Dens_S2 = mean(Dens_S2, na.rm = T),
      habval = mean(habval, na.rm = T)
    ) %>% 
    ungroup %>% 
    spread(habvar, habval) 
  
  return(dens2_avg)
  
})

# dens1_avg model
dens1_avgm <- reactive({
  
  # input
  dens1_avg <- dens1_avg() %>% 
    na.omit
  hbv <- input$hbv
  
  # formula
  frm <- paste0('log10(Dens_S1) ~ Year +', hbv) %>% 
    formula
  
  # global
  modoutglo <- lm(frm, data = dens1_avg, na.action = na.pass)
  modoutsel <-  modoutglo %>% 
    dredge %>% 
    get.models(subset = 1) %>% 
    .[[1]]
  
  # output list
  out <- list(
    global = modoutglo, 
    selected = modoutsel
  )
  
  return(out)

})

# dens2_avg model
dens2_avgm <- reactive({
  
  # input
  dens2_avg <- dens2_avg() %>% 
    na.omit
  hbv <- input$hbv
  
  # formula
  frm <- paste0('log10(Dens_S2) ~ Year +', hbv) %>% 
    formula
  
  # global
  modoutglo <- lm(frm, data = dens2_avg, na.action = na.pass)
  modoutsel <-  modoutglo %>% 
    dredge %>% 
    get.models(subset = 1) %>% 
    .[[1]]
  
  # output list
  out <- list(
    global = modoutglo, 
    selected = modoutsel
  )
  
  return(out)

})
```

## S1 year class

```{r}
renderPlot({
  
  # input
  hbv <- input$hbv
  dens1_avg <- dens1_avg()

  toplo1 <- dens1_avg %>% 
    gather('var', 'val', -Year)
  toplo2 <- dens1_avg %>% 
    na.omit
    
  # regressions    
  p1 <- ggplot(toplo1, aes(x = Year, y = val)) +
    geom_point(size = 5, pch = 21, aes(fill = Year)) + 
    stat_smooth(method = 'lm', se = FALSE, colour = 'black', linetype = 'dashed') + 
    facet_wrap(~var, scales = 'free_y', ncol = 1, strip.position = 'left') + 
    scale_x_continuous(breaks = seq(min(toplo2$Year), max(toplo2$Year))) +
    scale_fill_gradient(low = 'deepskyblue', high = 'lightgreen') +
    theme_bw(base_family = 'serif', base_size = 16) +
    theme(
      strip.placement = 'outside', 
      strip.background = element_blank(), 
      axis.title.y = element_blank(), 
      panel.grid = element_blank(),
      axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
      legend.position = 'none'
      )

  # time path
  p2 <- ggplot(toplo2, aes_string(x = hbv, y = 'Dens_S1')) + 
    geom_path(aes(colour = Year)) + 
    geom_label(aes(label = Year, fill = Year)) +
    scale_y_log10('log10-Density (fish/100ft)') +
    scale_fill_gradient(low = 'deepskyblue', high = 'lightgreen') +
    scale_colour_gradient(low = 'deepskyblue', high = 'lightgreen') +
    theme_bw(base_family = 'serif', base_size = 16) +
    theme(
      legend.position = 'none',
      panel.grid = element_blank()
      )

  grid.arrange(p1, p2, ncol = 2)
  
}, height = 500, width = 900)
```

<br>
Summary of global (left) and selected (right) linear model testing `r renderText({input$hbv})` and year effects, `r renderText({input$shd})` watershed, `r renderText({input$typ})` habitats.  See [here](images/model_table_explanations.pdf) for an explanation of the model result tables.

```{r, results = 'asis'}
renderUI({
  HTML(stargazer(dens1_avgm(), type = 'html', column.labels = c('Global', 'Selected'),
            dep.var.labels = 'S1 density (fish/100ft)'))
  })
```
<br>

## S2 year class

```{r}
renderPlot({
  
  # input
  hbv <- input$hbv
  dens2_avg <- dens2_avg()
  
  toplo1 <- dens2_avg %>% 
    gather('var', 'val', -Year)
  toplo2 <- dens2_avg %>% 
    na.omit
  
  # regressions    
  p1 <- ggplot(toplo1, aes(x = Year, y = val)) +
    geom_point(size = 5, pch = 21, aes(fill = Year)) + 
    stat_smooth(method = 'lm', se = FALSE, colour = 'black', linetype = 'dashed') + 
    facet_wrap(~var, scales = 'free_y', ncol = 1, strip.position = 'left') + 
    scale_x_continuous(breaks = seq(min(toplo2$Year), max(toplo2$Year))) +
    scale_fill_gradient(low = 'deepskyblue', high = 'lightgreen') +
    theme_bw(base_family = 'serif', base_size = 16) +
    theme(
      strip.placement = 'outside', 
      strip.background = element_blank(), 
      axis.title.y = element_blank(), 
      panel.grid = element_blank(),
      axis.text.x = element_text(size = 12, angle = 45, hjust = 1), 
      legend.position = 'none'
      )

  # time path
  p2 <- ggplot(toplo2, aes_string(x = hbv, y = 'Dens_S2')) + 
    geom_path(aes(colour = Year)) + 
    geom_label(aes(label = Year, fill = Year)) +
    scale_y_log10('log10-Density (fish/100ft)') +
    scale_fill_gradient(low = 'deepskyblue', high = 'lightgreen') +
    scale_colour_gradient(low = 'deepskyblue', high = 'lightgreen') +
    theme_bw(base_family = 'serif', base_size = 16) +
    theme(
      legend.position = 'none',
      panel.grid = element_blank()
      )

  grid.arrange(p1, p2, ncol = 2)
    
}, height = 500, width = 900)
  
```

<br>
Summary of global (left) and selected (right) linear model testing `r renderText({input$hbv})` and year effects, `r renderText({input$shd})` watershed, `r renderText({input$typ})` habitats.

```{r, results = 'asis'}
renderUI({
  HTML(stargazer(dens2_avgm(), type = 'html', column.labels = c('Global', 'Selected'),
            dep.var.labels = 'S2 density (fish/100ft)'))
  })
```
<br>

# Bivariate comparisons, summaries {.tabset}

Summary of all models in `r renderText({input$shd2})` watershed.  Only models where variables selected were other than or in addition to year are shown.

```{r}
column(12, 
       column(4, NULL),
       column(4, 
      selectInput('shd2', 'Select waterhsed:', sort(unique(dat$Watershed))) 
       ),
      column(4, NULL)
)
```

```{r}
# dens1modpr
allfctprs1 <- reactive({
  
  # inputs
  shd2 <- input$shd2
  
  tabout <- allfctprs %>% 
    filter(Watershed %in% shd2 & densvar %in% 'Dens_S1')
  
  return(tabout)
  
})

# dens2modpr
allfctprs2 <- reactive({
  
  # inputs
  shd2 <- input$shd2
  
  tabout <- allfctprs %>% 
    filter(Watershed %in% shd2 & densvar %in% 'Dens_S2')
  
  return(tabout)
  
})
```

## S1 year class {.tabset .tabset-fade .tabset-pills}

### pool

```{r, results = 'asis'}
renderUI({
  
  totab <- allfctprs1() %>% 
    filter(HabType %in% 'pool')
    
  if(nrow(totab) == 0)
    HTML('None')
  else 
    HTML(stargazer(totab$modsel, type = 'html',
            dep.var.labels = 'S1 density (fish/100ft)'))
  
  })
```
<br>

### riffle

```{r, results = 'asis'}
renderUI({
  
  totab <- allfctprs1() %>% 
    filter(HabType %in% 'riffle')
  
  if(nrow(totab) == 0)
    HTML('None')
  else 
    HTML(stargazer(totab$modsel, type = 'html',
            dep.var.labels = 'S1 density (fish/100ft)'))
  })
```
<br>

### run

```{r, results = 'asis'}
renderUI({
  
  totab <- allfctprs1() %>% 
    filter(HabType %in% 'run')
  
  if(nrow(totab) == 0)
    HTML('None')
  else 
    HTML(stargazer(totab$modsel, type = 'html',
            dep.var.labels = 'S1 density (fish/100ft)'))
  
  })
```
<br>

## S2 year class {.tabset .tabset-fade .tabset-pills}

### pool

```{r, results = 'asis'}
renderUI({
  
  totab <- allfctprs2() %>% 
    filter(HabType %in% 'pool')
  
  if(nrow(totab) == 0)
    HTML('None')
  else 
    HTML(stargazer(totab$modsel, type = 'html',
            dep.var.labels = 'S2 density (fish/100ft)'))
  
  })
```
<br>

### riffle

```{r, results = 'asis'}
renderUI({
  
  totab <- allfctprs2() %>% 
    filter(HabType %in% 'riffle')
  
  if(nrow(totab) == 0)
    HTML('None')
  else 
    HTML(stargazer(totab$modsel, type = 'html',
            dep.var.labels = 'S2 density (fish/100ft)'))
  
  })
```
<br>

### run

```{r, results = 'asis'}
renderUI({
  
  totab <- allfctprs2() %>% 
    filter(HabType %in% 'run')
  
  if(nrow(totab) == 0)
    HTML('None')
  else 
    HTML(stargazer(totab$modsel, type = 'html',
            dep.var.labels = 'S2 density (fish/100ft)'))
  
  })
```
<br>

# Model selection, all habitat variables {.tabset}

Models of salmonid density in relation to year and habitat variables are shown below.  Models were developed separately for each watershed and habitat type using all available data, where the station identifier was used as a random effect.  All combinations of habitat variables with a year effect from one to all were evaluated.  The top five models based on minimum AIC are shown.

```{r}
column(12, 
       column(4, NULL),
       column(4,
      selectInput('shd3', 'Select waterhsed:', sort(unique(dat$Watershed))) 
       ),
      column(4, NULL)
)
```

```{r}
# alltops, S1, selected shed
alltops1 <- reactive({
  
  # inputs
  shd3 <- input$shd3
  
  tabout <- alltops %>% 
    filter(Watershed %in% shd3 & densvar %in% 'Dens_S1')
  
  return(tabout)
  
})

# alltops, S2, selected shed
alltops2 <- reactive({
  
  # inputs
  shd3 <- input$shd3
  
  tabout <- alltops %>% 
    filter(Watershed %in% shd3 & densvar %in% 'Dens_S2')
  
  return(tabout)
  
})
```

## S1 year class {.tabset .tabset-fade .tabset-pills}

### pool

```{r, results = 'asis'}
renderUI({
  
  totab <- alltops1() %>% 
    filter(HabType %in% 'pool')

  HTML(stargazer(totab$mods, type = 'html',
          dep.var.labels = 'S1 density (fish/100ft)'))
  
  })
```
<br>

### riffle

```{r, results = 'asis'}
renderUI({
  
  totab <- alltops1() %>% 
    filter(HabType %in% 'riffle')

  HTML(stargazer(totab$mods, type = 'html',
          dep.var.labels = 'S1 density (fish/100ft)'))
  
  })
```
<br>

### run

```{r, results = 'asis'}
renderUI({
  
  totab <- alltops1() %>% 
    filter(HabType %in% 'run')

  HTML(stargazer(totab$mods, type = 'html',
          dep.var.labels = 'S1 density (fish/100ft)'))
  
  })
```
<br>

## S2 year class {.tabset .tabset-fade .tabset-pills}

### pool

```{r, results = 'asis'}
renderUI({
  
  totab <- alltops2() %>% 
    filter(HabType %in% 'pool')

  HTML(stargazer(totab$mods, type = 'html',
          dep.var.labels = 'S2 density (fish/100ft)'))
  
  })
```
<br>

### riffle

```{r, results = 'asis'}
renderUI({
  
  totab <- alltops2() %>% 
    filter(HabType %in% 'riffle')

  HTML(stargazer(totab$mods, type = 'html',
          dep.var.labels = 'S2 density (fish/100ft)'))
  
  })
```
<br>

### run

```{r, results = 'asis'}
renderUI({
  
  totab <- alltops2() %>% 
    filter(HabType %in% 'run')

  HTML(stargazer(totab$mods, type = 'html',
          dep.var.labels = 'S2 density (fish/100ft)'))
  
  })
```
<br>
